---
title: "DAA Taxa"
---

# Differential abundance analysis

```{r setup}
library(maaslin3)

source("../funct.R")
# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
taxa_levels <- c("species","genus", "family", "phylum")

prevalence.th <- 10/100
detection.th <- 0.1/100
formula_used_accross_gp <- '~ diet'
formula_used_within_gp <- '~ duration + (1 | id)'
formula_used_int <- "~ diet * duration + (1 | id)"
variable <- "diet" 
correction <- "Benjamini-Hochberg FDR adjustment"
```

Analysis information:

-   Method: maaslin3
-   Taxonomic level: `r taxa_levels`
-   Group variable: `r variable`
-   Formula used for interaction: **`r formula_used_int`**
-   Formula used across groups: **`r formula_used_accross_gp`**
-   Formula used within groups: **`r formula_used_within_gp`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r data_prep}
# Create output directory for results
output_base_dir <- "../../output/daa/daa_taxa"
dir.create(output_base_dir, showWarnings = FALSE)
```

```{r maaslin all}
# Run Maaslin3
run_maaslin3 <- function(taxa_tse, taxa_name, formula_used, dir) {
  # Keep only paired cases (individuals with both timepoint points)
  # paired.cases <- names(which(table(taxa_tse$subject) == 2))
  # taxa_tse <- taxa_tse[, taxa_tse$subject %in% paired.cases]
  # taxa_tse$treat <- taxa_tse$group

  # Run MaAsLin3
  fit <- maaslin3(
    taxa_tse, 
    output = dir,
    formula = formula_used,
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,
    standardize = TRUE,
    max_significance = 0.25,
    plot_associations = FALSE,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "WARN"
  )
  
  return(fit)
}
```

```{r}
file_path <- file.path("../../output/daa/daa_taxa/species/significant_results.tsv")
 if (!file.exists(file_path)) {
  lapply(taxa_levels, function(taxa_name) {
    taxa_tse <- altExp(tse, paste0(taxa_name, "_prevalent"))
    formula_used_int <- '~ diet * duration + (1 | id)'
    # Create output directory for this taxonomic level
    output_dir <- file.path(output_base_dir, taxa_name)
    dir.create(output_dir, showWarnings = FALSE)
    fit <- run_maaslin3(taxa_tse, taxa_name, formula_used_int, output_dir)
  
  })
}
```

## Interaction Model
```{r print_table}
#| label: tbl-table-of-significant-results
#| results: 'asis'
tabs <- NULL
tabs <- lapply(taxa_levels, function(taxa_name) {
  taxa_tse <- altExp(tse, paste0(taxa_name, "_prevalent"))
  total_features <- nrow(taxa_tse)
  file_path <- file.path("../../output/daa/daa_taxa", taxa_name, "all_results.tsv")
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of significant features for", taxa_name, 
                           "- Total features analyzed:", total_features)
    
  # Split the results by model
  daa_tab_split <- split(daa_tab, daa_tab$model)
  daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    rename( value = "term") %>%
      select(-model)
  
  # Format output for HTML or LaTeX
  datatable(
    daa_tab_split$abundance,
    options = list(pageLength = 50, dom = 'Bfrtip'),
    caption = table_caption,
    rownames = FALSE
  ) %>%
   formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
})
```

```{r}
tabs[[1]]
tabs[[2]]
tabs[[3]]
tabs[[4]]
```

## Post-Hoc Analysis

### Within-Group Comparisons
```{r run_maaslin_o}
taxa_tse <- altExp(tse, "species_prevalent")
tse_O <- taxa_tse[, taxa_tse$diet == "oat"]
tse_R <- taxa_tse[, taxa_tse$diet == "rice"]
taxa_name <- "species"

formula_used_int <- '~ duration + (1 | id)'
file_path <- file.path("../../output/daa/daa_taxa/oat/species/significant_results.tsv")
tse_list = list(tse_O, tse_R)
if (!file.exists(file_path)) {
  lapply(tse_list, function(tse){
    diet_dir <- file.path(output_base_dir, unique(tse$diet), taxa_name)
    dir.create(diet_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, taxa_name, formula_used_int, diet_dir)
  }
    )
}
```

```{r}
file_path_w <- file.path("../../data/wilcoxon_tables/wilcox_species.csv")
daa_tab_w <- read_csv(file_path_w, col_names = TRUE, show_col_types = FALSE)
daa_tab_w <- daa_tab_w %>% rename(Taxon = "feature")
```


```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$diet), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Comparison between timepoints within group", unique(tse$diet),"." , taxa_name, "level, relative abundance."," Method: Maaslin3, FDR: Wilcoxon", "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_w <- daa_tab_w[daa_tab_w$Diet == unique(tse$diet), ]  %>%
      select(-Diet)
    daa_tab_split$abundance <- merge(daa_tab_split$abundance, daa_tab_w, by = "feature") %>%
      select(-model)
    
    some_order <- c("feature", "mean.rel1",	"mean.rel2", "logFC.clr",	"coef",	"qval_individual",	"qval_joint",	"FDR")
      daa_tab_split$abundance <-  daa_tab_split$abundance  %>% 
        select(all_of(some_order))
      
      plot1 <- ggplot(daa_tab_split$abundance, aes(x = coef, y = logFC.clr)) +
        geom_point() +
        labs(title = "a) Internal validation") +
        geom_smooth(method = "lm", se = FALSE, color = "red")
      
      volcano_plot <- ggplot(daa_tab_split$abundance, aes(x = coef, y = qval_joint, size = mean.rel2)) +
        geom_point(aes(color = qval_joint < 0.25), alpha = 0.7) +  
        labs(
          title = "b) Volcano Plot: significance vs effect size",
          x = "Estimates (coef)",
          y = "qval_joint(<0.25)"
        ) +
        scale_color_manual(values = c("red", "blue")) +  
        theme_minimal() +
        theme(
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)
        )
      volcano_plot <- volcano_plot + 
        geom_text_repel(data = subset(daa_tab_split$abundance, qval_joint < 0.25),
                        aes(label = feature), size = 3, box.padding = 0.3)
      
      plot <- wrap_plots(plot1, volcano_plot)
  
    
      table <- datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 50, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
      
      list(plot = plot, table = table)
}
  )
```


```{r}
#| label: fig-figure-within-genus_all
#| fig-width: 8
#| fig-height: 8

tab_heat <- lapply(tse_list, function(tse){
    file_path <- file.path("../../output/daa/daa_taxa", unique(tse$diet), taxa_name, "all_results.tsv")
    df <- read_tsv(file_path, show_col_types = FALSE)
    df$Diet <- unique(tse$diet) 
    df_split <- split(df, df$model)
    return(df_split$abundance)
}
  )
tab_heat <- bind_rows(tab_heat)

tab_heat <- merge(tab_heat, daa_tab_w, by = c("feature", "Diet"))

tab_heat <- tab_heat %>%
  mutate(
    signif_label = case_when(
      qval_joint < 0.05 ~ "*",
      qval_joint < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(logFC.clr), paste0(round(logFC.clr, 4), signif_label), "")
  )

tab_heat <- tab_heat[tab_heat$qval_joint < 0.25, ]

# Plot
ggplot(tab_heat, aes(x = Diet, y = feature, fill = logFC.clr)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Effect Size"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

#### Diet Oat - Estimates the change in relative abundance between timepoint 1 and timepoint 2
```{r}
#| label: tbl-table-within-speciesO
tabs[[1]]$table
```

```{r}
#| label: fig-figure-within-speciesO
#| fig-cap: "Diet Oat: (a)  Internal validation: checking that alternative tests give correlated results, (b) volcano plot showing the relation between effect size and significance"
#| fig-width: 10
#| fig-height: 5
tabs[[1]]$plot
```

#### Diet Rice - Estimates the change in relative abundance between timepoint 1 and timepoint 2
```{r}
#| label: tbl-table-within-speciesB
tabs[[2]]$table
```

```{r}
#| label: fig-figure-within-speciesB
#| fig-cap: "Diet Rice (a)  Internal validation: checking that alternative tests give correlated results, (b) volcano plot showing the relation between effect size and significance"
#| fig-width: 10
#| fig-height: 5
tabs[[2]]$plot
```

### Across-Group Comparisons
```{r}
tse_after <- taxa_tse[, taxa_tse$timepoint == "after"]
tse_before <- taxa_tse[, taxa_tse$timepoint == "before"]

formula_used_int <- '~ diet'

tse_list = list(tse_after, tse_before)
if (!file.exists(file_path)) {
lapply(tse_list, function(tse){
  timepoint_dir <- file.path(output_base_dir, unique(tse$timepoint), taxa_name)
  dir.create(timepoint_dir, showWarnings = FALSE, recursive = TRUE)
  
  run_maaslin3(tse, taxa_name, formula_used_int, timepoint_dir)
}
  )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$timepoint), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Comparison between between treatments(reference = 4) at timepoint", unique(tse$duration), "." , taxa_name, "level, relative abundance."," Method: Maaslin3",
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    # rename( value = "term") %>%
      select(-model)
  
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 50, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

#### Before Intervention - Estimates the difference in relative abundance between treatments at week 0
```{r}
#| label: tbl-table-before-species
tabs[[1]]
```

#### Post Intervention  Estimates the difference in relative abundance between treatments at week 6
```{r}
#| label: tbl-table-after-species
tabs[[2]]
```


```{r setup_m}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
tse <- tse[, tse$paired=="yes"]
taxa_levels <- c("species")

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Diet" 
formula_used <- "~ diet * timepoint + (1 | id)"
correction <- "Benjamini-Hochberg FDR adjustment"

```

## The following box plots highlight significant species from the interaction model.

This section highlights significant taxa from maaslin3.

Analysis information:

-   Method: maaslin3
-   Taxonomic level: `r taxa_levels`
-   Group variable: `r variable`
-   Formula used: **`r formula_used`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r tsesub_m}
file_path1 <- file.path("../../output/daa/daa_taxa/species/significant_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
file_path3 <- file.path("../../output/daa/daa_taxa/oat/species/significant_results.tsv")
file_path4 <- file.path("../../output/daa/daa_taxa/rice/species/significant_results.tsv")
daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE)
daa_tab3 <- read_tsv(file_path3, col_names = TRUE, show_col_types = FALSE)
daa_tab4 <- read_tsv(file_path4, col_names = TRUE, show_col_types = FALSE)
significant_species <- unique(c(daa_tab1$feature, daa_tab3$feature, daa_tab4$feature))

daa_tab <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% significant_species) %>%
      select(feature,metadata, value, qval_joint) %>%
      mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r before-after_m}
tse$diet <- factor(tse$diet, levels=c("oat", "rice"))

```

```{r}

taxa_tse <- altExp(tse, "species_prevalent")
df <- mia::meltSE(taxa_tse, add.col = T, assay.type = "relabundance")

df_not_zer0 <- df %>%
    group_by(FeatureID, duration, diet) %>%
    summarise(
        nonzero_n = sum(relabundance > 0, na.rm = TRUE),
        total_n = sum(!is.na(relabundance)),
        pct_nonzero = round(100 * nonzero_n / total_n, 1),
        .groups = "drop"
    ) %>%
    filter(nonzero_n > 0) %>%
    pivot_wider(
        names_from = c(duration, diet),
        values_from = c(nonzero_n, total_n, pct_nonzero),
        names_glue = "{diet}_{duration}_{.value}"
    )

plots_list <- list()
plots_list <- setNames(lapply(significant_species, function(taxon_name) {
    
    p1 <- plotBoxplot(taxa_tse, assay.type = "relabundance", x = "diet",  fill.by = "duration", pair.by = "id", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = T, ncol = 1, label.size = 5) +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )

    p2 <- plotBoxplot(taxa_tse, assay.type = "relabundance", x = "duration",  fill.by = "diet", features = c(taxon_name), facet.by = "rownames", add.significance = T, ncol = 1, label.size = 5) +
    scale_fill_viridis_d(option = "D") +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )
    
    label_text <- generate_prevalence_label(df_not_zer0, taxon_name)
    
    p12 <- wrap_plots(p1, p2, axes = "collect") + 
      labs(caption = label_text) +
      theme(
    plot.caption = element_text(size = 10, 
                                face = "bold.italic", 
                                hjust = 1, color = "darkblue") ) 
    
    return(p12)
    
}
    ),
 significant_species)

```



```{r hits_m}
#| label: fig-figure-of-highlights
#| layout-ncol: 1
#| fig-width: 10 
#| fig-height: 6 

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```
