---
title: "Differential abundance analysis"
author: ""
params: 
  tax.level: default
---

### Differential abundance analysis

```{r setup}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Define variables and formulas
variable <- "diet" 
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- params$tax.level

# Add CLR transformation
tse <- transformAssay(
  altExp(tse, tax.level), assay.type = "relabundance",
  method = "clr",
  pseudocount=TRUE
)
```

Analysis information:

-   Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r detection.th`
-   Prevalence threshold: `r prevalence.th`
-   Number of taxonomic groups: `r nrow(tse)`

```{r tsesub}
# pick tse for time point 1 and 2 separately
tse <- tse[, tse$paired=="yes"]
tse1 <- tse[, tse$timepoint=="before"]
tse2 <- tse[, tse$timepoint=="after"]

```

```{r within_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (tp in unique(tse$timepoint)) {
  # Pick data for the given time point
  tse2 <- tse[, tse$timepoint==tp]
  pvs <- apply(assay(tse2, "clr"), 1, function (x) {kruskal.test(x ~ tse2$diet)$p.value})
  padj <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$timepoint)
pvals.timepoints <- pvals
```

Number of significant findings within time points (Kruskal test, Benjamini-Hochberg FDR adjustment).

```{r between_timepoint_analysis2}
print(colSums(pvals.timepoints<0.25))
```

```{r between_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (g in unique(tse$diet)) {
  # Pick data for the given Treatment for time points 1 and 2
  tse1  <- tse[, tse$diet==g & tse$timepoint=="before"]
  tse2  <- tse[, tse$diet==g & tse$timepoint=="after"]
  # Ensure that subject match (pairing ok)
  if (!all(tse1$id==tse2$id)) {stop("Subjects do not match!")}
  pvs   <- sapply(rownames(tse), function (tax) {
               wilcox.test(assay(tse1, "clr")[tax,], assay(tse2, "clr")[tax,], paired=TRUE)$p.value
	     }
	   )
  padj  <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$diet)
pvals.treatments <- pvals
```

Number of significant findings within treatments (Paired Wilcoxon test, Benjamini-Hochberg FDR adjustment).

```{r between_timepoint_analysis3}
print(colSums(pvals.treatments<0.25))
```

Let us look more closely to the significant hits. The logFC corresponds to logFC in CLR (tp2-tp1).

```{r hits}
#| layout-ncol: 3
#| fig-width: 10
#| fig-height: 8
library(reshape2)
theme_set(theme_bw(30))
tabs <- NULL # Collect results in a table
for (g in unique(tse$diet)) {

  # P-values
  signif.features <- names(which(pvals.treatments[, g]<0.25))

  if (length(signif.features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in signif.features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]
      
      df <- df %>%
        filter(paired == "yes") %>%
        group_by(id) %>%
        mutate(
            change = if_else(time == 2, .data[["signal"]] - .data[["signal"]][time == 1], 0),
            change = if_else(time == 1, change[time == 2], change)
        ) %>%
        ungroup() %>%
        as.data.frame(check.names = FALSE)

      tabs <-
        rbind(tabs,
        c(
          Group = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
      
      p <- ggplot(df, aes(x=timepoint, y=signal)) +
             geom_boxplot(outlier.shape=NA) +
             # geom_jitter(width=0.05) +
             geom_point() + 	     	     
             geom_line(aes(group = id, color = change)) +
          scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab") +
	     scale_y_log10(labels = scales::percent) + 
	     labs(title=paste(g, "/", tax),
	          y="Abundance (%)")  
      print(p)
    }
  }
}

# Clean up the table
tabs <- as.data.frame(tabs)
tabs$mean.relabundance1 <- round(100 * as.numeric(tabs$mean1), 3); tabs$mean1 <- NULL
tabs$mean.relabundance2 <- round(100 * as.numeric(tabs$mean2), 3); tabs$mean2 <- NULL
tabs$FDR <- round(as.numeric(tabs$FDR), 4)
tabs$logFC.clr <- round(as.numeric(tabs$logFC.clr), 4)
tabs <- tabs %>%
  arrange(FDR, desc(logFC.clr))
```

For a complete table of results, see [complete results](https://seafile.utu.fi/f/a5d57b061c3249e193da/)

```{r table}
#| label: table
if (knitr::is_html_output()) {
  datatable(tabs,
            options = list(pageLength = 10),
            caption = "Table of significant hits")
} else {
  kable(tabs)
}
```

```{r volcano-tab}
tabs_all <- NULL
for (g in unique(tse$diet)) {
  # P-values
  all_features <-  names(pvals.treatments[, g])

  if (length(all_features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in all_features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars], verbose = FALSE)
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      tabs_all <-
        rbind(tabs_all,
        c(
          Group = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          # Standard Deviations
        # sd1 = sd(subset(df, timepoint == "before")$signal),
        # sd2 = sd(subset(df, timepoint == "after")$signal),
        # 
        # # Medians
        # median1 = median(subset(df, timepoint == "before")$signal),
        # median2 = median(subset(df, timepoint == "after")$signal),
        # 
        # # IQR (25th and 75th percentiles)
        # iqr25_1 = quantile(subset(df, timepoint == "before")$signal, 0.25),
        # iqr75_1 = quantile(subset(df, timepoint == "before")$signal, 0.75),
        # 
        # iqr25_2 = quantile(subset(df, timepoint == "after")$signal, 0.25),
        # iqr75_2 = quantile(subset(df, timepoint == "after")$signal, 0.75),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
    }
  }
}
```


```{r volcano-tab-fix}
tabs_all <- as.data.frame(tabs_all)
tabs_all$mean.rel1 <- round(100 * as.numeric(tabs_all$mean1), 1); tabs_all$mean1 <- NULL
tabs_all$mean.rel2 <- round(100 * as.numeric(tabs_all$mean2), 1); tabs_all$mean2 <- NULL
tabs_all$FDR <- round(as.numeric(tabs_all$FDR), 3)
tabs_all$logFC.clr <- round(as.numeric(tabs_all$logFC.clr), 4)
```

```{r volcano-plot}
volcano_plot <- ggplot(tabs_all, aes(x = logFC.clr, y = FDR, size = mean.rel2)) +
  geom_point(aes(color = FDR < 0.25), alpha = 0.7) +  
  # geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") + 
  # geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") + 
  # scale_x_log10() +
  labs(
    title = "Volcano Plot: Effect Size vs Significance",
    x = "Log Fold Change (CLR)",
    y = "FDR"
  ) +
  scale_color_manual(values = c("gray", "blue")) +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
volcano_plot + 
  geom_text_repel(data = subset(tabs_all, FDR < 0.25),
                  aes(label = Taxon), size = 3, box.padding = 0.3)
```


