---
title: "DAA Taxa"
---

# Differential abundance analysis

```{r setup}
library(maaslin3)

source("../funct.R")
# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
taxa_levels <- c("species","genus", "family", "phylum")

prevalence.th <- 10/100
detection.th <- 0.1/100
formula_used_accross_gp <- '~ diet'
formula_used_within_gp <- '~ duration + (1 | id)'
formula_used_int <- "~ diet * duration + (1 | id)"
variable <- "diet" 
correction <- "Benjamini-Hochberg FDR adjustment"
```

Analysis information:

-   Method: maaslin3
-   Taxonomic level: `r taxa_levels`
-   Group variable: `r variable`
-   Formula used for interaction: **`r formula_used_int`**
-   Formula used across groups: **`r formula_used_accross_gp`**
-   Formula used within groups: **`r formula_used_within_gp`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r data_prep}
# Create output directory for results
output_base_dir <- "../../output/daa/daa_taxa"
dir.create(output_base_dir, showWarnings = FALSE)
```

```{r maaslin all}
# Run Maaslin3
run_maaslin3 <- function(taxa_tse, taxa_name, formula_used, dir) {
  # Keep only paired cases (individuals with both timepoint points)
  # paired.cases <- names(which(table(taxa_tse$subject) == 2))
  # taxa_tse <- taxa_tse[, taxa_tse$subject %in% paired.cases]
  # taxa_tse$treat <- taxa_tse$group

  # Run MaAsLin3
  fit <- maaslin3(
    taxa_tse, 
    output = dir,
    formula = formula_used,
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,
    standardize = TRUE,
    max_significance = 0.25,
    plot_associations = FALSE,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "WARN"
  )
  
  return(fit)
}
```

```{r}
file_path <- file.path("../../output/daa/daa_taxa/species/significant_results.tsv")
 if (!file.exists(file_path)) {
  lapply(taxa_levels, function(taxa_name) {
    taxa_tse <- altExp(tse, paste0(taxa_name, "_prevalent"))
    formula_used_int <- '~ diet * duration + (1 | id)'
    # Create output directory for this taxonomic level
    output_dir <- file.path(output_base_dir, taxa_name)
    dir.create(output_dir, showWarnings = FALSE)
    fit <- run_maaslin3(taxa_tse, taxa_name, formula_used_int, output_dir)
  
  })
}
```


```{r print_table}
#| label: table-of-significant-results
#| results: 'asis'
tabs <- NULL
tabs <- lapply(taxa_levels, function(taxa_name) {
  taxa_tse <- altExp(tse, paste0(taxa_name, "_prevalent"))
  total_features <- nrow(taxa_tse)
  file_path <- file.path("../../output/daa/daa_taxa", taxa_name, "all_results.tsv")
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of significant features for", taxa_name, 
                           "- Total features analyzed:", total_features)
    
  # Split the results by model
  daa_tab_split <- split(daa_tab, daa_tab$model)
  
  # Format output for HTML or LaTeX
  datatable(
    daa_tab_split$abundance,
    options = list(pageLength = 6, dom = 'Bfrtip'),
    caption = table_caption,
    rownames = FALSE
  ) %>%
   formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
})
```

```{r}
tabs[[1]]
tabs[[2]]
tabs[[3]]
tabs[[4]]
```


### Post-Hoc Analysis

## Within-Group Comparisons
```{r run_maaslin_o}
taxa_tse <- altExp(tse, "species_prevalent")
tse_O <- taxa_tse[, taxa_tse$diet == "oat"]
tse_R <- taxa_tse[, taxa_tse$diet == "rice"]
taxa_name <- "species"

formula_used_int <- '~ duration + (1 | id)'
file_path <- file.path("../../output/daa/daa_taxa/oat/species/significant_results.tsv")
tse_list = list(tse_O, tse_R)
if (!file.exists(file_path)) {
  lapply(tse_list, function(tse){
    diet_dir <- file.path(output_base_dir, unique(tse$diet), taxa_name)
    dir.create(diet_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, taxa_name, formula_used_int, diet_dir)
  }
    )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$diet), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for", taxa_name, "within group", unique(tse$diet),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    
  
    
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 6, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Diet Oat
```{r}
tabs[[1]]
```

### Diet Rice
```{r}
tabs[[2]]
```

## Across-Group Comparisons
```{r}
tse_after <- taxa_tse[, taxa_tse$timepoint == "after"]
tse_before <- taxa_tse[, taxa_tse$timepoint == "before"]

formula_used_int <- '~ diet'

tse_list = list(tse_after, tse_before)
if (!file.exists(file_path)) {
lapply(tse_list, function(tse){
  timepoint_dir <- file.path(output_base_dir, unique(tse$timepoint), taxa_name)
  dir.create(timepoint_dir, showWarnings = FALSE, recursive = TRUE)
  
  run_maaslin3(tse, taxa_name, formula_used_int, timepoint_dir)
}
  )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$timepoint), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for", taxa_name, "at", unique(tse$duration),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
  
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 6, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Before Intervention - Week 0
```{r}
tabs[[1]]
```

### Post Intervention - Week 6
```{r}
tabs[[2]]
```


```{r setup_m}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
tse <- tse[, tse$paired=="yes"]
taxa_levels <- c("species")

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Diet" 
formula_used <- "~ diet * timepoint + (1 | id)"
correction <- "Benjamini-Hochberg FDR adjustment"

```

## Maaslin3 significant hits

This section highlights significant taxa from maaslin3.

Analysis information:

-   Method: maaslin3
-   Taxonomic level: `r taxa_levels`
-   Group variable: `r variable`
-   Formula used: **`r formula_used`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r tsesub_m}
file_path1 <- file.path("../../output/daa/daa_taxa/species/significant_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE)
significant_specise <- unique(daa_tab1$feature)

daa_tab <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% significant_specise) %>%
      select(feature,metadata, value, qval_joint) %>%
      mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r before-after_m}
tse$diet <- factor(tse$diet, levels=c("oat", "rice"))
df <- as.data.frame(colData(tse))

plot_taxon <- function(taxon) {
  
  # Check if the taxon exists in genus or species
  if (taxon %in% rowData(tse)$genus | taxon %in% rowData(tse)$species) {
    
    # Calculate signal (relative abundance)
        if (taxon %in% rowData(tse)$genus) {
          rownames(tse) <- rowData(tse)$genus
          df$signal <- as.vector(assay(tse, "relabundance")[taxon,])
        } else {
          rownames(tse) <- rowData(tse)$species
          df$signal <- as.vector(assay(tse, "relabundance")[taxon,])
        }
    
    daa_tab_tax <- daa_tab %>%
      filter(feature == taxon)
    
    anot_tax <- daa_tab_tax %>% 
        filter(metadata == "intervention")
    
    anot_tax$pval_color <- ifelse(anot_tax$qval_joint <= 0.25, "red", "blue")
    # Subset for before and after
    df_before <- df %>% filter(timepoint == "before")
    df_after <- df %>% filter(timepoint == "after")
    
    # Prepare data for plotting
        vars <- c("id", "timepoint", "signal")
        dfm <- melt(df[, vars])
        colnames(dfm) <- vars
        
        # Filter and mutate for paired design
        df <- df %>%
          filter(paired == "yes") %>%
          group_by(id) %>%
          mutate(
            change = if_else(timepoint == "after", signal - signal[timepoint == "before"], 0),
            change = if_else(timepoint == "before", change[timepoint == "after"], change)
          ) %>%
          ungroup() %>%
          as.data.frame(check.names = FALSE)
        
          min_change <- min(df$change, na.rm = TRUE)
          max_change <- max(df$change, na.rm = TRUE)
          median_change <- median(df$change, na.rm = TRUE)
    
    p12 <- ggplot(df, aes(x = diet, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 15), legend.position = "none") +
      labs(title = taxon, y = "Abundance (%)", x = "Diet") +
      facet_wrap(~ duration, nrow = 1)
    
    p3 <- ggplot(df, aes(x = duration, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      geom_line(aes(group = id, color = change)) +
      scale_color_gradient2(midpoint = median_change, low = "blue", mid = "white", high = "red", space = "Lab", limits = c(min_change, max_change)) +
      facet_wrap(~ diet) +
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 15), legend.position = "right") +
      labs(x = "Time", y = "Abundance (%)")
    
    # Combine the before and after plots
    return(wrap_plots(p12, p3, axes = "collect")
  )
  }

}
```

```{r}
plots_list <- list()
plots_list <- setNames(lapply(significant_specise, plot_taxon),
                      significant_specise)
```

```{r hits_m}
#| layout-ncol: 1
#| fig-width: 10 
#| fig-height: 3 

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```
