---
title: "DAA Metabolic Pathways"
---

# Interaction Model

```{r setup}
library(maaslin3)

source("../funct.R")
# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

pa_tse <- altExp(tse, "pathabundance")

variable <- "diet" 
formula_used <- "~ diet * timepoint + (1 | id)"
```

-   Group variable: **`r variable`**
-   Formula used: **`r formula_used`**

```{r data_prep}
# Create output directory for results
output_base_dir <- "../../output/daa/daa_pa"
dir.create(output_base_dir, showWarnings = FALSE)
# Keep paired cases that have both time points
paired.cases <- names(which(table(colData(pa_tse)$id) == 2))
pa_tse <- pa_tse[, pa_tse$id %in% paired.cases]
```

```{r maaslin all}
# Run Maaslin3
run_maaslin3 <- function(taxa_tse, formula_used, dir) {
  # Keep only paired cases (individuals with both timepoint points)
  # paired.cases <- names(which(table(taxa_tse$subject) == 2))
  # taxa_tse <- taxa_tse[, taxa_tse$subject %in% paired.cases]
  # taxa_tse$treat <- taxa_tse$group

  # Run MaAsLin3
  fit <- maaslin3(
    taxa_tse, 
    output = dir,
    formula = formula_used,
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,
    standardize = TRUE,
    max_significance = 0.25,
    plot_associations = FALSE,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "WARN"
  )
  
  return(fit)
}
```

```{r}
file_path <- file.path("../../output/daa/daa_pa/interaction/significant_results.tsv")
 if (!file.exists(file_path)) {
    formula_used_int <- '~ diet * duration + (1 | id)'
    # Create output directory for this fxnomic level
    output_dir <- file.path(output_base_dir, "interaction")
    dir.create(output_dir, showWarnings = FALSE)
    fit <- run_maaslin3(pa_tse, formula_used_int, output_dir)
}
```

```{r print_table}
#| label: table-of-significant-results
file_path <- file.path("../../output/daa/daa_pa/interaction/all_results.tsv")
total_features <- nrow(pa_tse)

daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
    filter(!is.na(feature)) %>%
    select(feature, metadata, value, coef, qval_individual, qval_joint, model)

table_caption <- paste("Table of Metabolic Pathways- Total features analyzed:", total_features)
  
# Split the results by model
daa_tab_split <- split(daa_tab, daa_tab$model)
daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    rename(metadata = "variable", value = "term") %>%
      select(-model)

# Format output for HTML or LaTeX
datatable(
  daa_tab_split$abundance,
  options = list(pageLength = 50, dom = 'Bfrtip'),
  caption = table_caption,
  rownames = FALSE
) %>%
 formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)

```

# Post-Hoc Analysis

## Within-Group Comparisons

```{r run_maaslin_o}
tse_O <- pa_tse[, pa_tse$diet == "oat"]
tse_R <- pa_tse[, pa_tse$diet == "rice"]

formula_used_int <- '~ duration + (1 | id)'
file_path <- file.path("../../output/daa/daa_pa/oat/significant_results.tsv")
tse_list = list(tse_O, tse_R)
if (!file.exists(file_path)) {
  lapply(tse_list, function(tse){
    diet_dir <- file.path(output_base_dir, unique(tse$diet))
    dir.create(diet_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, formula_used_int, diet_dir)
  }
    )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_pa", unique(tse$diet), "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for path coverage within group", unique(tse$diet),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
      select(c(-model, -metadata, -value))
    
  
    
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 50, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Diet Oat

```{r}
tabs[[1]]
```

### Diet Rice

```{r}
tabs[[2]]
```

## Across-Group Comparisons

```{r}
tse_after <- pa_tse[, pa_tse$timepoint == "after"]
tse_before <- pa_tse[, pa_tse$timepoint == "before"]

formula_used_int <- '~ diet'
file_path <- file.path("../../output/daa/daa_pa", unique(tse_before$timepoint), "all_results.tsv")
tse_list = list(tse_after, tse_before)
if (!file.exists(file_path)) {
lapply(tse_list, function(tse){
  timepoint_dir <- file.path(output_base_dir, unique(tse$timepoint))
  dir.create(timepoint_dir, showWarnings = FALSE, recursive = TRUE)
  
  run_maaslin3(tse, formula_used_int, timepoint_dir)
}
  )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_pa", unique(tse$timepoint), "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for metacyc at", unique(tse$duration),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
      select(c(-model, -metadata, -value))
  
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 50, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Before Intervention - Week 0

```{r}
tabs[[1]]
```

### Post Intervention - Week 6

```{r}
tabs[[2]]
```

```{r setup_m}

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Diet" 
formula_used <- "~ diet * timepoint + (1 | id)"
correction <- "Benjamini-Hochberg FDR adjustment"

```

## Maaslin3 significant hits

This section highlights significant taxa from maaslin3.

Analysis information:

-   Method: maaslin3
-   Group variable: `r variable`
-   Formula used: **`r formula_used`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r tsesub_m}
file_path1 <- file.path("../../output/daa/daa_pa/interaction/significant_results.tsv")
file_path2 <- file.path("../../output/daa/daa_pa/oat/significant_results.tsv")
file_path3 <- file.path("../../output/daa/daa_pa/rice/significant_results.tsv")
daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE)
daa_tab2 <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) 
daa_tab3 <- read_tsv(file_path3, col_names = TRUE, show_col_types = FALSE) 
daa_tab1 <- na.omit(daa_tab1)
daa_tab2 <- na.omit(daa_tab2)
daa_tab3 <- na.omit(daa_tab3)
significant_fxns <- unique(c(daa_tab1$feature, daa_tab2$feature, daa_tab3$feature))
target_list <- rownames(pa_tse)

closest_matches <- sapply(significant_fxns, function(x) {
  target_list[which.min(stringdist(x, target_list, method = "jw"))]
})


```

```{r before-after_m}
pa_tse$diet <- factor(pa_tse$diet, levels=c("oat", "rice"))
df <- mia::meltSE(pa_tse, add.col = T, assay.type = "relabundance")

df_not_zer0 <- df %>%
    group_by(FeatureID, duration, diet) %>%
    summarise(
        nonzero_n = sum(relabundance > 0, na.rm = TRUE),
        total_n = sum(!is.na(relabundance)),
        pct_nonzero = round(100 * nonzero_n / total_n, 1),
        .groups = "drop"
    ) %>%
    filter(nonzero_n > 0) %>%
    pivot_wider(
        names_from = c(duration, diet),
        values_from = c(nonzero_n, total_n, pct_nonzero),
        names_glue = "{diet}_{duration}_{.value}"
    )

```

```{r}

plots_list <- list()
plots_list <- setNames(lapply(closest_matches, function(taxon_name) {
    
    p1 <- plotBoxplot( pa_tse, assay.type = "relabundance", x = "diet",  fill.by = "duration", pair.by = "id", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = T, ncol = 1, label.size = 5) +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15),
        strip.text = element_blank(),
        axis.text = element_text(face = "bold"),
  ) + labs(title = taxon_name)

    p2 <- plotBoxplot( pa_tse, assay.type = "relabundance", x = "duration",  fill.by = "diet", features = c(taxon_name), facet.by = "rownames", add.significance = T, ncol = 1, label.size = 5) +
        scale_fill_viridis_d(option = "D") +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        strip.text = element_blank(),
        axis.text = element_text(face = "bold"),
  )
    
    label_text <- generate_prevalence_label(df_not_zer0, taxon_name)
    
    p12 <- wrap_plots(p1, p2, axes = "collect") + 
      labs(caption = label_text) +
      theme(
    plot.caption = element_text(size = 10, 
                                face = "bold.italic", 
                                hjust = 1, color = "darkblue") )
    
    return(p12)
    
}
    ),
 closest_matches)
```

```{r hits_m}
#| layout-ncol: 1
#| fig-width: 10 
#| fig-height: 6 

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```
