---
title: "Differential abundance analysis"
author: ""
---

### Differential abundance analysis

```{r setup}
# Define variables and formulas
variable <- "diet" 
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- "genus"

# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Add CLR transformation
tse <- transformAssay(
  altExp(tse, "genus_prevalent"), assay.type = "relabundance",
  method = "clr",
  pseudocount=1
)
```

Analysis information:

-   Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r prevalence.th`
-   Number of taxonomic groups: `r nrow(tse)`

BMI is not significantly different between conditions or time points. Hence we can ignore it in the differential abundance testing. In group "B" it seems to be borderline significant (p\<0.05 before multiple testing adjustment).

```{r bmicheck}
# Check whether BMI is significantly different between conditions:
# Between conditions within time point 1
# Between conditions within time point 2
# Within conditions between time points (paired analysis)

# pvs <- c()
# for (tp in unique(tse$timepoint)) {
#   tse.sub <- tse[, tse$timepoint==tp]
#   pvs[[tp]] <- kruskal.test(tse.sub$BMI ~ tse.sub$Treatment)$p.value
# }

# pick tse for time point 1 and 2 separately
tse <- tse[, tse$paired=="yes"]
tse1 <- tse[, tse$timepoint=="before"]
tse2 <- tse[, tse$timepoint=="after"]
# Paired wilcoxon test for BMI within each treatment
# for (g in unique(tse$Treatment)) {
#   # subject order matches
#   if (!all(tse1[, tse1$Treatment==g]$subject==tse2[, tse2$Treatment==g]$subject)) {stop("Subjects do not match")}
# 
#   pvs[[g]] <- wilcox.test(tse1[, tse1$Treatment==g]$BMI,
#                           tse2[, tse2$Treatment==g]$BMI,
# 	                  paired=TRUE)$p.value
# }
# padj <- p.adjust(pvs, method="bonferroni")
# print(padj)
```

```{r within_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (tp in unique(tse$timepoint)) {
  # Pick data for the given time point
  tse2 <- tse[, tse$timepoint==tp]
  pvs <- apply(assay(tse2, "clr"), 1, function (x) {kruskal.test(x ~ tse2$diet)$p.value})
  padj <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$timepoint)
pvals.timepoints <- pvals
```

Number of significant findings within time points.

-   Kruskal test, Benjamini-Hochberg FDR adjustment

```{r between_timepoint_analysis2}
print(colSums(pvals.timepoints<0.25))
```

```{r between_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (g in unique(tse$diet)) {
  # Pick data for the given Treatment for time points 1 and 2
  tse1  <- tse[, tse$diet==g & tse$timepoint=="before"]
  tse2  <- tse[, tse$diet==g & tse$timepoint=="after"]
  # Ensure that subject match (pairing ok)
  if (!all(tse1$id==tse2$id)) {stop("Subjects do not match!")}
  pvs   <- sapply(rownames(tse), function (tax) {
               wilcox.test(assay(tse1, "clr")[tax,], assay(tse2, "clr")[tax,], paired=TRUE)$p.value
	     }
	   )
  padj  <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$diet)
pvals.treatments <- pvals
```

Number of significant findings within treatments:

-   Paired Wilcoxon test, Benjamini-Hochberg FDR adjustment

```{r between_timepoint_analysis3}
print(colSums(pvals.treatments<0.25))
```

Let us look more closely to the significant hits. The logFC corresponds to logFC in CLR (tp2-tp1).

```{r hits}
#| layout-ncol: 3
#| fig-width: 10
#| fig-height: 10
library(reshape2)
theme_set(theme_bw(30))
tabs <- NULL # Collect results in a table
for (g in unique(tse$diet)) {

  # P-values
  signif.features <- names(which(pvals.treatments[, g]<0.25))

  # Pick data for the given Treatment for time points 1 and 2
  #tse1  <- tse[, tse$Treatment==g & tse$timepoint=="before"]
  #tse2  <- tse[, tse$Treatment==g & tse$timepoint=="after"]

  if (length(signif.features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- colData(tseg)
    for (tax in signif.features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      tabs <-
        rbind(tabs,
        c(
          Group = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
      
      p <- ggplot(df, aes(x=timepoint, y=signal)) +
             geom_boxplot(outlier.shape=NA) +
             # geom_jitter(width=0.05) +
             geom_point() + 	     	     
             geom_line(aes(group=id), color="gray") +
	     scale_y_log10() + 
	     labs(title=paste(g, "/", tax),
	          y="Abundance (%)")
      print(p)
    }
  }
}

# Clean up the table
tabs <- as.data.frame(tabs)
tabs$mean.relabundance1 <- round(100 * as.numeric(tabs$mean1), 3); tabs$mean1 <- NULL
tabs$mean.relabundance2 <- round(100 * as.numeric(tabs$mean2), 3); tabs$mean2 <- NULL
tabs$FDR <- round(as.numeric(tabs$FDR), 4)
tabs$logFC.clr <- round(as.numeric(tabs$logFC.clr), 4)
```

```{r table}
library(knitr)
kable(tabs)
```
