---
title: "Differential abundance analysis"
author: ""
params: 
  tax.level: default
---

```{r setup}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Define variables and formulas
variable <- "diet" 
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- params$tax.level

# Add CLR transformation
tse <- transformAssay(
  altExp(tse, tax.level), assay.type = "relabundance",
  method = "clr",
  pseudocount=TRUE
)
```

Analysis information:

-   Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%
-   Number of taxonomic groups: `r nrow(tse)`

```{r tsesub}
# pick tse for time point 1 and 2 separately
tse <- tse[, tse$paired=="yes"]
tse1 <- tse[, tse$timepoint=="before"]
tse2 <- tse[, tse$timepoint=="after"]

```

```{r within_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (tp in unique(tse$timepoint)) {
  # Pick data for the given time point
  tse2 <- tse[, tse$timepoint==tp]
  pvs <- apply(assay(tse2, "clr"), 1, function (x) {kruskal.test(x ~ tse2$diet)$p.value})
  padj <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$timepoint)
pvals.timepoints <- pvals
```

Number of significant findings within time points (Kruskal test, Benjamini-Hochberg FDR adjustment).

```{r between_timepoint_analysis2}
print(colSums(pvals.timepoints<0.25))
```

```{r between_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (g in unique(tse$diet)) {
  # Pick data for the given Treatment for time points 1 and 2
  tse1  <- tse[, tse$diet==g & tse$timepoint=="before"]
  tse2  <- tse[, tse$diet==g & tse$timepoint=="after"]
  # Ensure that subject match (pairing ok)
  if (!all(tse1$id==tse2$id)) {stop("Subjects do not match!")}
  pvs   <- sapply(rownames(tse), function (tax) {
               wilcox.test(assay(tse1, "clr")[tax,], assay(tse2, "clr")[tax,], paired=TRUE)$p.value
	     }
	   )
  padj  <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$diet)
pvals.treatments <- pvals
```

Number of significant findings within treatments (Paired Wilcoxon test, Benjamini-Hochberg FDR adjustment).

```{r between_timepoint_analysis3}
print(colSums(pvals.treatments<0.25))
```

Let us look more closely to the significant hits. The logFC corresponds to logFC in CLR (tp2-tp1).

```{r tsefix}
file_path1 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/genus/all_results.tsv")


signif.features <- names(which(pvals.treatments[, 1] < 0.25 | pvals.treatments[, 2] < 0.25))

taxa_of_interest <- c(signif.features)

daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint) 
daa_tab2 <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint)

daa_tab <- rbind(daa_tab1, daa_tab2) %>%
  mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r before-after}
df <- as.data.frame(colData(tse))

plot_taxon <- function(taxon) {
  
  # Check if the taxon exists in genus or species
  if (taxon %in% rowData(tse)$genus | taxon %in% rowData(tse)$species) {
    
    # Calculate signal (relative abundance)
        if (taxon %in% rowData(tse)$genus) {
          rownames(tse) <- rowData(tse)$genus
          df$signal <- as.vector(assay(tse, "relabundance")[taxon,])
        } else {
          rownames(tse) <- rowData(tse)$species
          df$signal <- as.vector(assay(tse, "relabundance")[taxon,])
        }
    
    daa_tab_tax <- daa_tab %>%
      filter(feature == taxon)
    
    anot_tax <- daa_tab_tax %>% 
        filter(value == "time")
    
    anot_tax$pval_color <- ifelse(anot_tax$qval_joint <= 0.25, "red", "blue")
    # Subset for before and after
    df_before <- df %>% filter(timepoint == "before")
    df_after <- df %>% filter(timepoint == "after")
    
    # Generate "Before treatment" plot
    p1 <- ggplot(df_before, aes(x = diet, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 40), legend.position = "none") +
      # ggpubr::stat_pvalue_manual(
      #   data = final_results_before %>% filter(taxon == !!taxon),
      #   xmin = "group1",
      #   xmax = "group2",
      #   y.position = "y.position",
      #   label = "p.adj"
      # ) +
      labs(title = "Before", x = "Diet", y = "Abundance (%)")
    
    # Generate "After treatment" plot
    p2 <- ggplot(df_after, aes(x = diet, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 40), legend.position = "none") +
      # ggpubr::stat_pvalue_manual(
      #   data = final_results_after %>% filter(taxon == !!taxon),
      #   xmin = "group1",
      #   xmax = "group2",
      #   y.position = "y.position",
      #   label = "p.adj"
      # ) +
      labs(title = "After", x = "Diet", y = "Abundance (%)")
    
    # Combine the before and after plots
    return(wrap_plots(p1, p2, axes = "collect") + 
                plot_annotation(
    title = taxon,
    theme = theme(plot.title = ggtext::element_markdown())
  )
  # plot_annotation(
  #   title = paste(taxon, "/", "p =", ifelse(taxon %in% anot_tax$feature, 
  #                        paste0("<span style='color:", anot_tax$pval_color[anot_tax$feature == taxon], "'>", 
  #                               anot_tax$qval_joint[anot_tax$feature == taxon], "</span>"), 
  #                        "<span style='color:gray;'>NA</span>")
  #   ),
  #   theme = theme(plot.title = ggtext::element_markdown())
  # )
  )
  }
  
  return(NULL)
}
```

```{r}
plots_list <- list()
plots_list <- setNames(lapply(taxa_of_interest, plot_taxon),
                       taxa_of_interest)
```


```{r hits}
#| layout-ncol: 3
#| fig-width: 10
#| fig-height: 8
library(reshape2)
theme_set(theme_bw(30))
tabs <- NULL 

for (g in unique(tse$diet)) {

  if (length(signif.features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in signif.features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]
      
      df <- df %>%
        filter(paired == "yes") %>%
        group_by(id) %>%
        mutate(
            change = if_else(time == 2, .data[["signal"]] - .data[["signal"]][time == 1], 0),
            change = if_else(time == 1, change[time == 2], change)
        ) %>%
        ungroup() %>%
        as.data.frame(check.names = FALSE)
      
      min_change <- min(df$change, na.rm = TRUE)
      max_change <- max(df$change, na.rm = TRUE)
      median_change <- median(df$change, na.rm = TRUE)
      
      daa_tab_tax <- daa_tab %>%
            filter(feature == tax)
          
      anot_tax <- daa_tab_tax %>% 
          filter(value == g)
      
      anot_tax$pval_color <- ifelse(anot_tax$qval_joint <= 0.25, "red", "blue")

      tabs <-
        rbind(tabs,
        c(
          Group = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
      
      p <- ggplot(df, aes(x=timepoint, y=signal)) +
             geom_boxplot(outlier.shape=NA) +
             # geom_jitter(width=0.05) +
             geom_point() + 	     	     
             geom_line(aes(group = id, color = change)) +
          scale_color_gradient2(midpoint=median_change, 
                                low="blue", 
                                mid="white", 
                                high="red", 
                                space ="Lab",
                                limits = c(min_change, max_change)) +
	     scale_y_log10(labels = scales::percent) + 
          labs(title = g, y = "Abundance (%)") +
# 	     labs(
#             title = paste0(g, "/", " p = ", 
#                            ifelse(tax %in% anot_tax$feature, 
#                                   paste0("<span style='color:", anot_tax$pval_color[anot_tax$feature == tax], "'>", 
#                                          anot_tax$qval_joint[anot_tax$feature == tax], "</span>"), 
#                                   "<span style='color:gray;'>NA</span>")),
#             y = "Abundance (%)"
#           ) +
          theme(
            plot.title = ggtext::element_markdown() 
          )  
      
      plots_list[[paste(tax, "0", g, sep = "_")]] <- p
    }
  }
}

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```


```{r}
# Clean up the table
tabs <- as.data.frame(tabs)
tabs$mean.rel1 <- round(100 * as.numeric(tabs$mean1), 3); tabs$mean1 <- NULL
tabs$mean.rel2 <- round(100 * as.numeric(tabs$mean2), 3); tabs$mean2 <- NULL
tabs$FDR <- round(as.numeric(tabs$FDR), 4)
tabs$logFC.clr <- round(as.numeric(tabs$logFC.clr), 4)
tabs <- tabs %>%
  arrange(FDR, desc(logFC.clr))
```


For a complete table of results, see [complete results](https://seafile.utu.fi/f/a5d57b061c3249e193da/)

```{r table}
#| label: table
if (knitr::is_html_output()) {
  datatable(tabs,
            options = list(pageLength = 10),
            caption = "Table of significant hits")
} else {
  kable(tabs)
}
```

```{r volcano-tab}
tabs_all <- NULL
for (g in unique(tse$diet)) {
  # P-values
  all_features <-  names(pvals.treatments[, g])

  if (length(all_features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in all_features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars], verbose = FALSE)
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      tabs_all <-
        rbind(tabs_all,
        c(
          Group = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          # Standard Deviations
        # sd1 = sd(subset(df, timepoint == "before")$signal),
        # sd2 = sd(subset(df, timepoint == "after")$signal),
        # 
        # # Medians
        # median1 = median(subset(df, timepoint == "before")$signal),
        # median2 = median(subset(df, timepoint == "after")$signal),
        # 
        # # IQR (25th and 75th percentiles)
        # iqr25_1 = quantile(subset(df, timepoint == "before")$signal, 0.25),
        # iqr75_1 = quantile(subset(df, timepoint == "before")$signal, 0.75),
        # 
        # iqr25_2 = quantile(subset(df, timepoint == "after")$signal, 0.25),
        # iqr75_2 = quantile(subset(df, timepoint == "after")$signal, 0.75),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
    }
  }
}
```


```{r volcano-tab-fix}
tabs_all <- as.data.frame(tabs_all)
tabs_all$mean.rel1 <- round(100 * as.numeric(tabs_all$mean1), 1); tabs_all$mean1 <- NULL
tabs_all$mean.rel2 <- round(100 * as.numeric(tabs_all$mean2), 1); tabs_all$mean2 <- NULL
tabs_all$FDR <- round(as.numeric(tabs_all$FDR), 3)
tabs_all$logFC.clr <- round(as.numeric(tabs_all$logFC.clr), 4)
```

```{r volcano-plot}
#| fig.width: 12
#| fig.height: 8

volcano_plot <- ggplot(tabs_all, aes(x = logFC.clr, y = FDR, size = mean.rel2)) +
  geom_point(aes(color = FDR < 0.25), alpha = 0.7) +  
  # geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") + 
  # geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") + 
  # scale_x_log10() +
  labs(
    title = "Volcano Plot: Effect Size vs Significance",
    x = "Log Fold Change (CLR)",
    y = "FDR"
  ) +
  scale_color_manual(values = c("blue", "red")) +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
volcano_plot + 
  geom_text_repel(data = subset(tabs_all, FDR < 0.25),
                  aes(label = Taxon), size = 3, box.padding = 0.3)
```


