---
title: "Beta diversity"
format:
  html:
    code-fold: true
    number-sections: true
    page-layout: full
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

## Community composition

This section reports on a microbiota project results of community composition across different diets and visits.

**Groups**

```{r data}
source("funct.R")
tse <- readRDS("../data/tse.Rds")
```

### Phylum

```{r community-composition-phyla}
#| fig-width: 20
#| fig-height: 9
meals <- c(
  "1" = "1",
  "2" = "2",
  "3" = "3",
  "4" = "4"
)
# Visualization - composition bar plot
plots <- lapply(names(meals), function(meal_id) {
    label <- meals[[meal_id]]
    # Filter tse for the current meal group
    tse_filtered <- tse[, tse$meal_group == meal_id]
    # Plot abundance for the filtered tse
    plotAbundance(
        altExp(tse_filtered, "phylum"),
        assay.type = "relabundance",
        color.by = "phylum_sub",
        group = "phylum",
        order.row.by = "abund"
    ) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        labs(title = label, x = "Phylum", y = "Relative Abundance")
})
bar_plot <- wrap_plots(plots, ncol = 2, guides = "collect")
bar_plot
```

### Genus

```{r community-composition-genus}
#| fig-width: 20
#| fig-height: 9
# Visualization - composition bar plot
plots <- lapply(names(meals), function(meal_id) {
    label <- meals[[meal_id]]
    # Filter tse for the current meal group
    tse_filtered <- tse[, tse$meal_group == meal_id]
    # Plot abundance for the filtered tse
    plotAbundance(
      altExp(tse_filtered, "genus"),
      assay.type = "relabundance",
      color.by = "genus_sub", 
      group = "genus", 
      order.row.by = "abund"
    ) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        labs(title = label)
})
bar_plot <- wrap_plots(plots, ncol = 2, guides = "collect")
bar_plot
```

### Heatmap

```{r heatmap}
#| fig.width: 20
#| fig.height: 10
#| out.width: "100%"
#| fig.align: "center"

tse_heat <- transformAssay(
    altExp(tse, "phylum"), assay.type = "counts", method = "relabundance", pseudocount = 1)
tse_heat <- transformAssay(tse_heat, assay.type = "relabundance", method = "clr")
tse_heat <- transformAssay(
    tse_heat, assay.type = "clr", MARGIN = "rows", method = "standardize",
    name = "clr_z")
# Plot heatMap with sechm
plots <- lapply(names(meals), function(meal_id) {
    label <- meals[[meal_id]]
    # Filter tse for the current meal group
    tse_heat_filtered <- tse_heat[, tse_heat$meal_group == meal_id]
    heatmap <- sechm(
        tse_heat_filtered, assayName = "clr_z", features=rownames(tse_heat_filtered),
        show_rownames = TRUE, show_colnames = TRUE,
        annocol_title_side = "left", do.scale=TRUE,
        row_names_gp = gpar(fontsize = 6), column_names_gp = gpar(fontsize = 8),
        breaks = 1, anno_colors = list("blue", "red", "green", "purple"),
        width = unit(7, "inches"),
        height = unit(3, "inches")
        )
    heatmap <- grid.grabExpr(draw(heatmap))
})
grid.newpage()
do.call(grid.arrange, c(plots, ncol = 2))
```

## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# Define variables
tax.level <- "species"
dissimilarity.index <- "bray"
variable <- "group"  
```

**Taxonomic level: `r tax.level`**

**Dissimilarity index: `r dissimilarity.index`**

**Group variable: `r variable`**

PCoA plots

```{r pcoa-plot, fig.show="hold"}
tse <- runMDS(
        tse,
        FUN = getDissimilarity,
        method = "bray",
        assay.type = "counts",
        name = "PCoA_BC"
    )
# Calculate explained variance
e <- attr(reducedDim(tse, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])
    
# Create the PCoA plot
p <- plotReducedDim(tse, "PCoA_BC", 
                    colour_by = variable, 
                    other_fields = c("meal_group", "group"))
# Add explained variance for axes
p1 <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
) + facet_wrap(~meal_group, nrow = 2, labeller = labeller(meal_group = meals))
# Add ellipses for groups
p_ellipse <- p1 + stat_ellipse(aes(color = tse$group), level = 0.95) +
  scale_color_manual(
    values = c("blue", "red", "green", "purple"), 
    ) +
  theme(
      legend.position = "bottom",
      legend.box.margin = margin(-10, 0, 0, 0)
  ) +
  coord_fixed(ratio = 1)

p_ellipse 
```

### Significant variables

```{r significant variables}
# Perform RDA
tse <- addRDA(
    tse,
    assay.type = "relabundance",
    formula = assay ~ meal_group + timepoint + age + gender,
    distance = dissimilarity.index,
    na.action = na.exclude)

# Calculate explained variance
e_rda <- attr(reducedDim(tse, "RDA"), "eig")
rel_eig_rda <- e / sum(e[e > 0])
# Generate RDA plot coloured by Treatment
plotRDA(tse, "RDA", colour.by = "meal_group") + 
  labs(
    x = paste("RDA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
    y = paste("RDA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
  ) +
  scale_color_manual(
    values = c("blue", "red", "green", "purple"), 
    labels = c("Oat-Rice", "Rice-Oat", "Rice-Rice", "Oat-Oat")
    ) + 
  labs(color = "Meal Group") +
  theme(legend.position = "top", legend.title = element_text(size = 12), legend.text = element_text(size = 10))
```

### Significant taxa

```{r significant-taxa}
plotLoadings(tse, "RDA", ncomponents = 2, n = 20)
```

## Supervised analysis

### Permanova analysis

```{r run_rdba_tests}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova <- function(tse, variable, comparison) {
    tse_subset <- tse[, colData(tse)$group %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = X ~ group, 
                          permutations = 999)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res)
    group_row <- perm_df[rownames(perm_df) == "group", ]
    
    result <- data.frame(
        Group1 = comparison[1],
        Group2 = comparison[2],
        P_value = group_row$"permanova.Pr..F.",
        Homogeneity = signif(tail(perm_df$homogeneity.Explained.variance, 1), 3),
        P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests <- function(tse, comparisons, variable) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova(tse, variable, comp)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_before-after}
# For the treatment comparisons
final_results_permanova <- run_rdba_tests(tse, comparisons, variable)
```

```{r permanova_res}
#| label: permanova_res
if (knitr::is_html_output()) {
  datatable(final_results_permanova,
            options = list(pageLength = 4),
            caption = "PERMANOVA Results - All Comparisons")
} else {
  library(kableExtra)
  final_results_permanova %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - All Comparisons",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova)-1))) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows(index = table(final_results_permanova$Time),
             indent = FALSE,
             hline_after = TRUE)
}
```

## Alternate analysis

```{r data-load-again}
tse <- readRDS("../data/tse.Rds")
```

```{r early-late}
# Early/Late analysis
beta_res <- lapply(seq(2), function(i, timepoint) {
    tse_sub <- tse[,colData(tse) %>%
                      as_tibble(rownames="sample_id") %>%
                      filter(timepoint %in% timepoint[[i]]) %>%
                      tidyr::drop_na(c("group", "age")) %>%
                      select(sample_id) %>% .[[1]]]
    
    set.seed(123)
    betadiso_pval <- anova(betadisper(
        vegdist(t(assay(tse_sub, "relabundance")), method = "bray"),
        colData(tse_sub)[,"group"]
    ))["Groups","Pr(>F)"]
    
    mat <- t(assay(tse_sub, "relabundance"))
    set.seed(123)
    permanova_pval <- adonis2(
        as.formula(paste0("mat ~ ", paste0(c("group", "meal_group"), collapse = "+"))),
        data = colData(tse_sub),
        method = "bray", 
        permutations = 999,
        strata = NULL)["Model", "Pr(>F)"]
    
    c(betadisp_pval = betadiso_pval,
      permanova_pval = permanova_pval)
}, timepoint = list("before", "after")) %>%
    setNames(c("before", "after")) %>%
    as.data.frame()
```

```{r change}
# Change analysis
beta_res_change <- lapply(
    levels(colData(tse)[,"meal_group"]),
    function(lv) {
        tse_sub <- tse[,colData(tse) %>%
                          as_tibble(rownames="sample_id") %>%
                          filter(paired == "yes") %>%
                          tidyr::drop_na(c("group", "age")) %>%
                          filter(meal_group == lv) %>%
                          select(sample_id) %>% .[[1]]]
        
        mat <- t(assay(tse_sub, "relabundance"))
        set.seed(123)
        permanova_pval <- adonis2(
            as.formula(paste0("mat ~ ", paste0(c("timepoint", "group"), collapse = "+"))),
            data = colData(tse_sub), 
            method = "bray", 
            permutations = 999,
            strata = NULL)["Model", "Pr(>F)"]
        permanova_pval
    }) %>% 
    setNames(paste0("change-", levels(colData(tse)[,"meal_group"]))) %>% 
    data.frame(row.names = c("permanova_pval"), check.names = FALSE)

beta_res_change["betadisp_pval",] = rep(NA, nlevels(colData(tse)[,"meal_group"]))
beta_res <- merge(beta_res, beta_res_change, by="row.names")
rownames(beta_res) <- beta_res[,"Row.names"]
beta_res <- beta_res[,-1]
```

```{r viz, fig.height= 8, fig.width= 8}
# Visualization
set.seed(123)
tse <- runMDS(tse, FUN = vegan::vegdist, method = "bray",
              name = "PCoA_BC", exprs_values = "relabundance")
e <- attr(reducedDim(tse, "PCoA_BC"), "eig")
rel_eig <- e/sum(e[e>0])

# Plot 1: PCoA by Treatment
p1 <- plotReducedDim(tse[,!is.na(colData(tse)[["group"]])], "PCoA_BC",
                     colour_by = "group",
                     other_fields = c("meal_group", "group")) +
    facet_wrap(~meal_group, ncol = 2) +
    labs(x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
         y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = ""),
         title = "") +
    theme(legend.position = "bottom", 
          legend.direction = "horizontal",
          plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(name = "group", 
                      values = colorRampPalette(c("blue", "red"))(nlevels(colData(tse)[["group"]]))) +
    stat_ellipse(aes(color = group))

# Plot 2: Trajectories
tse <- getStepwiseDivergence(tse, group = "group",
                            time_field = "time",
                            assay_name = "relabundance")

p2 <- plotReducedDim(tse[,!is.na(colData(tse)[["group"]])],
                     "PCoA_BC", colour_by = "timepoint",
                     other_fields = c("group", "id", "meal_group", "timepoint")) +
    aes(color = timepoint) +
    labs(
        title = "Trajectories between meals",
        x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
        y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
    ) +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(
        values = c("before" = "blue", "after" = "orange"),
        name = "Timepoint"
    ) +
    facet_wrap(~meal_group, nrow = 1) +
    geom_line(
        aes(group = id),
        color = "grey50",
        alpha = 0.5,
        linetype = "dashed"
    )

# Plot 3: Divergence comparison
p3 <- colData(tse) %>% 
    as_tibble() %>%
    select(c("group", "time_divergence")) %>% 
    tidyr::drop_na(c("group", "time_divergence")) %>% 
    ggplot(aes(x = group, y = time_divergence)) + 
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.1, size = 0.5) +
    geom_signif(comparisons = combn(levels(colData(tse)[,"group"]), 2, simplify = FALSE),
                map_signif_level = FALSE, step_increase = 0.2, textsize = 2.5) +
    labs(title = "Comparing divergences per group") +
    theme(plot.title = element_text(hjust = 0.5))
```

```{r viz-fig1, fig.height=6, fig.width=8}
p1
```

```{r viz-fig2, fig.height=6, fig.width=12}
p2
```

```{r viz-fig3, fig.height=8, fig.width=8}
p3
```

```{r viz-tab}
#| label: beta_results
if (knitr::is_html_output()) {
datatable(beta_res,
          options = list(
            pageLength = 4, 
            dom = 'Bfrtip'    
          )) %>%
  formatStyle(
    columns = names(beta_res),  
    backgroundColor = styleInterval(
      cuts = 0.05,
      values = c('green', 'white')
    ),
    valueColumns = names(beta_res)  
  ) %>%
  formatRound(
    columns = names(beta_res),
    digits = 3
  ) 
} else {
  library(kableExtra)
  beta_res %>%
    as.data.frame() %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(beta_res)))) %>%
    row_spec(0, bold = TRUE) %>%
    kable_styling(latex_options = c("striped", "hold_position"))
}
```

### Data Summary

```{r data-summary}
tse
```

## Version info

```{r sessioninfo}
sessionInfo()
```
