---
title: "Beta diversity"
format:
  html:
    code-fold: true
    number-sections: true
    figures:
      layout: "grid"
      ncol: 2
      nrow: 2
editor: visual
execute:
  echo: true
---

## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# lapply(comparisons, function(comp) PCoA_plot(tse, comp,variable, outdir))
# final_results_permanova <- run_rdba_tests(tse, comparisons, variable, outdir)

# Define variables
tax.level <- "species"
dissimilarity.index <- "Bray-Curtis"
variable <- "group"  
```

**Taxonomic level: `r tax.level`**

**Dissimilarity index: `r dissimilarity.index`**

**Group variable: `r variable`**

```{r message=FALSE, warning=FALSE, echo=FALSE}
# Load and preprocess data
# tse <- readRDS("../data/tse.Rds")
source("data.R")
```

PCoA plots

```{r, fig.show="hold", message=FALSE, warning=FALSE, echo=FALSE}
# Generate plots
beta_pcoa <- lapply(comparisons, function(comp) PCoA_plot(altExp(tse, tax.level), comp, variable))
 combined_plot <- wrap_plots(beta_pcoa, nrow = 2)
 combined_plot
```

## Supervised analysis

### dbRDA and permanova analysis

```{r message=FALSE, echo=FALSE}
library(DT)

final_results_permanova <- run_rdba_tests(tse, comparisons, variable)
l <- names(final_results_permanova)
datatable(final_results_permanova[[l[[1]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[1]]))

datatable(final_results_permanova[[l[[2]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[2]]))

datatable(final_results_permanova[[l[[3]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[3]]))

datatable(final_results_permanova[[l[[4]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[4]]))

```
