---
title: "Beta diversity"
format:
  html:
    code-fold: true
    number-sections: true
    figures:
      layout: "grid"
      ncol: 2
      nrow: 2
editor: visual
execute:
  echo: true
---

## Community composition

This section reports on a microbiota project results of community composition across different diets and visits.

**Groups**

```{r data-load, message=FALSE, warning=FALSE, echo=FALSE}
# tse <- readRDS("../data/tse.Rds")
source("data.R")
```

## Data Summary

```{r data-summary, message=FALSE, warning=FALSE}
tse
```

## Community composition at the phyla level

```{r community-composition-phyla, message=FALSE, warning=FALSE, fig.height= 9, fig.width= 20, echo=FALSE}

# Visualization - composition bar plot
plotAbundance(
  tse_phylum,
  assay.type = "relabundance",
  color_by = "phylum_sub", group = "phylum", order.row.by = "abund"
) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Community composition at the genus level

```{r community-composition-genus, message=FALSE, warning=FALSE, fig.height= 9, fig.width= 20, echo=FALSE}

# Visualization - composition bar plot
plotAbundance(
  tse_genus,
  assay.type = "relabundance",
  color_by = "genus_sub", group = "genus", order.row.by = "abund") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# Define variables
tax.level <- "species"
dissimilarity.index <- "Bray-Curtis"
variable <- "group"  
```

**Taxonomic level: `r tax.level`**

**Dissimilarity index: `r dissimilarity.index`**

**Group variable: `r variable`**

PCoA plots

```{r, fig.show="hold", message=FALSE, warning=FALSE, echo=FALSE}
# Generate plots
beta_pcoa <- PCoA_plot(altExp(tse, tax.level), comparisons, variable)
 #combined_plot <- wrap_plots(beta_pcoa, nrow = 2)
 #combined_plot
beta_pcoa
```

## Supervised analysis

### dbRDA and permanova analysis

```{r message=FALSE, echo=FALSE}

final_results_permanova <- run_rdba_tests(tse, comparisons, variable)
l <- names(final_results_permanova)
datatable(final_results_permanova[[l[[1]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[1]]))

datatable(final_results_permanova[[l[[2]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[2]]))

datatable(final_results_permanova[[l[[3]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[3]]))

datatable(final_results_permanova[[l[[4]]]], 
        options = list(pageLength = 10), 
        caption = paste("Results for Group:", l[[4]]))

```
