---
title: "Beta diversity"
format: html
editor: visual
output-dir: ../output
execute:
  echo: true
---
## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta 
diversity analysis. Specifically, in this section,  we employed Principal 
Coordinates Analysis (PCoA) using the Bray-Curtis dissimilarity index to assess c
ommunity dissimilarity among microbiota samples.

Groups

```{r message=FALSE, warning = FALSE, echo=FALSE}
# Load and preprocess data
# tse <- readRDS("../data/tse.Rds")
source("funct.R")
source("data.R")
tse <- agglomerateByRanks(tse)
tse <- transformAssay(tse, method = "relabundance")
tse <- altExp(tse, "species")
```

PCoA plots

```{r message=FALSE, warning = FALSE, echo=FALSE}
# Generate plots
beta_pcoa <- lapply(comparisons, function(comp) PCoA_plot(tse, comp, variable))
combined_plot <- wrap_plots(beta_pcoa, nrow = 2)
combined_plot
```

## Supervised analysis

### dbRDA and permanova analysis

```{r message=FALSE, echo=FALSE}
library(DT)

final_results_permanova <- run_rdba_tests(tse, comparisons, variable)

datatable(final_results_permanova, options = list(pageLength = 10), caption = "PERMANOVA Results for Different Comparisons")
```

### Comparison of the microbial community composition between groups

```{r echo=FALSE, warning = FALSE, message=FALSE}
library(ggpubr)
library(tidyverse)


index <- "shannon"
group_var <- "group"

# Changes old levels with new levels
tse$group <- factor(tse$group)

# Calculate p values
pvals <- pairwise.wilcox.test(
    tse[[index]], tse[[group_var]], p.adjust.method = "fdr")
# Put them to data.frame format
pvals <- pvals[["p.value"]] |>
    as.data.frame()
varname <- "group1"
pvals[[varname]] <- rownames(pvals)
# To long format
pvals <- reshape(
    pvals,
    direction = "long",
    varying = colnames(pvals)[ !colnames(pvals) %in% varname ],
    times = colnames(pvals)[ !colnames(pvals) %in% varname ],
    v.names = "p",
    timevar = "group2",
    idvar = "group1"
    ) |>
    na.omit()
# Add y-axis position
pvals[["y.position"]] <- apply(pvals, 1, function(x){
    temp1 <- tse[[index]][ tse[[group_var]] == x[["group1"]] ]
    temp2 <- tse[[index]][ tse[[group_var]] == x[["group2"]] ]
    temp <- max( c(temp1, temp2) )
    return(temp)
})
pvals[["y.position"]] <- max(pvals[["y.position"]]) +
    order(pvals[["y.position"]]) * 0.2
# Round values
pvals[["p"]] <- round(pvals[["p"]], 3)

# Create a boxplot
p <- plotColData(
    tse, x = group_var, y = index,
    show_boxplot = TRUE, show_violin = FALSE) +
    theme(text = element_text(size = 10)) +
    stat_pvalue_manual(pvals)
p
```

