---
title: "Unsupervised analysis"
format: html
editor: visual
execute:
  echo: true
---

## Community dissimilarity

This document presents the results of the OGB microbiota project, focusing on beta diversity analysis. Specifically, we employed Principal Coordinates Analysis (PCoA) using the Bray-Curtis dissimilarity index to assess community dissimilarity among microbiota samples.

```{r data-load, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)     
library(TreeSummarizedExperiment)  
library(mia)
library(miaViz)
library(ggpubr)        
library(scater)        
```

Data Preprocessing

```{r}
# Load and preprocess data
tse <- readRDS("../data/tse.Rds")
tse <- agglomerateByRanks(tse)
tse <- transformAssay(tse, method = "relabundance")
tse <- altExp(tse, "species")
```

Define Comparisons

```{r}
# Define comparisons for subsetting
comparisons <- list(
  c("diet_1_visit_1", "diet_1_visit_2"),
  c("diet_2_visit_1", "diet_2_visit_2"),
  c("diet_1_visit_1", "diet_2_visit_1"),
  c("diet_1_visit_2", "diet_2_visit_2")
)

variable <- "group"
```

Subset and Analyze Data

### Plotting Function

```{r}
# Plotting function
# Perform PCoA
.plot_PCOA <- function(tse, comparison, variable) {
  # Subset the TSE to include only samples for the specified groups in the comparison                                                   
  tse_subset <- tse[, tse$group %in% comparison]
  tse_subset <- runMDS(
    tse_subset,
    FUN = getDissimilarity,
    method = "bray",
    assay.type = "counts",
    name = "PCoA_BC"
  )
  # Create ggplot object
  p <- plotReducedDim(tse_subset, "PCoA_BC", colour_by = variable)
  # Calculate explained variance
  e <- attr(reducedDim(tse_subset, "PCoA_BC"), "eig")
  rel_eig <- e / sum(e[e > 0])
  # Add explained variance for each axis
  p1 <- p + labs(
    title = paste("PCoA for Comparison", paste(comparison, collapse = " vs ")),
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = "")
  )
  p1
  p_ellipse <- p1 + stat_ellipse(aes(color = colour_by), level = 0.95)
  p_ellipse
}
```

Generate plots

```{r}
# Generate plots
lapply(comparisons, function(comp) .plot_PCOA(tse, comp, variable))
```
