---
title: "Community composition and Alpha Diversity Analysis (OGB microbiota project)"
author: "Nitin Bayal"
format: html
editor: visual
execute:
  echo: true
---

## Introduction

This document reports on the OGB microbiota project results of community composition and alpha diversity analyses across different diets and visits.

## Setup

```{r data-load, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(TreeSummarizedExperiment)
library(mia)

tse <- readRDS("../data/tse.Rds")
tse <- addAlpha(x = tse,assay.type = 'counts',index = c('shannon', 'simpson'))
```

## Data Summary

```{r data-summary, message=FALSE, warning=FALSE}
tse
```

The `echo: false` option disables the printing of code (only output is displayed).

## Community composition

```{r community-composition, message=FALSE, warning=FALSE, fig.height= 9, fig.width= 15}
# Load required libraries
library(mia)
library(miaViz)

# Load TreeSummarizedExperiment object
tse <- readRDS("../data/tse.Rds")

# Computing relative abundance
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Getting top taxa on a Genus level
tse <- agglomerateByRank(tse, rank = "genus")
top_genera <- getTop(tse, top = 10, assay.type = "relabundance")

# Renaming the "Genus" rank to keep only top taxa and the rest to "Other"
genus_renamed <- lapply(rowData(tse)$genus, function(x) {
    if (x %in% top_genera) { x } else { "Other" }
})
rowData(tse)$genus_sub <- as.character(genus_renamed)

# Agglomerate the data based on specified taxa
tse_genus <- agglomerateByVariable(tse, by = "rows", f = "genus_sub")

# Visualization - composition bar plot
plotAbundance(
    tse_genus, assay.type = "relabundance",
    order.row.by = "abund", facet.by = "group")
```

## Alpha Diversity: Observed Species

```{r observed-species, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 8}
# Load required libraries
library(mia)

# Load TreeSummarizedExperiment object
tse <- readRDS("../data/tse.Rds")

# Estimate (observed species) richness
tse <- addAlpha(x = tse, assay.type = "counts", index = c("simpson", "shannon"))

# Visualization - colData variables
library(scater)
plotColData(
    tse,
    "observed",
    "group",
    colour_by = "group") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(x = "groups", y = expression(Richness[Observed]))

```

### Group-wise comparisons using Shannon index

```{r group-comparisons-shannon, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 8}
# Load required libraries
library(mia)
library(TreeSummarizedExperiment)
library(ggpubr)
library(tidyverse)

# Define index and grouping variable
index <- "shannon"
group_var <- "group"

# Ensure the grouping variable is a factor
colData(tse)[[group_var]] <- factor(colData(tse)[[group_var]])

# Calculate p-values
pvals <- pairwise.wilcox.test(
  x = colData(tse)[[index]],
  g = colData(tse)[[group_var]],
  p.adjust.method = "fdr"
)

# Convert p-values to a data frame for ggpubr
pvals_df <- as.data.frame(pvals$p.value)
pvals_df$group1 <- rownames(pvals_df)
pvals_long <- pivot_longer(pvals_df, -group1, names_to = "group2", values_to = "p") %>%
  na.omit()

# Add y-axis position
pvals_long <- pvals_long %>%
  mutate(
    y.position = sapply(1:nrow(.), function(i) {
      max(c(
        colData(tse)[[index]][colData(tse)[[group_var]] == .$group1[i]],
        colData(tse)[[index]][colData(tse)[[group_var]] == .$group2[i]]
      ))
    })
  )
pvals_long$y.position <- max(pvals_long$y.position) + seq_len(nrow(pvals_long)) * 0.4
pvals_long$p <- round(pvals_long$p, 3)

# Visualization - boxplot
p <- plotColData(
  tse, x = group_var, y = index,
  show_boxplot = TRUE, show_violin = FALSE
) +
  theme(text = element_text(size = 10)) +
  stat_pvalue_manual(pvals_long, label = "p")

print(p)
```

### Group-wise comparisons using Simpson index

```{r group-comparisons-simpson, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 8}
# Load required libraries
library(mia)
library(TreeSummarizedExperiment)
library(ggpubr)
library(tidyverse)

# Define index and grouping variable
index <- "simpson"
group_var <- "group"

# Ensure the grouping variable is a factor
colData(tse)[[group_var]] <- factor(colData(tse)[[group_var]])

# Calculate p-values
pvals <- pairwise.wilcox.test(
  x = colData(tse)[[index]],
  g = colData(tse)[[group_var]],
  p.adjust.method = "fdr"
)

# Convert p-values to a data frame for ggpubr
pvals_df <- as.data.frame(pvals$p.value)
pvals_df$group1 <- rownames(pvals_df)
pvals_long <- pivot_longer(pvals_df, -group1, names_to = "group2", values_to = "p") %>%
  na.omit()

# Add y-axis position
pvals_long <- pvals_long %>%
  mutate(
    y.position = sapply(1:nrow(.), function(i) {
      max(c(
        colData(tse)[[index]][colData(tse)[[group_var]] == .$group1[i]],
        colData(tse)[[index]][colData(tse)[[group_var]] == .$group2[i]]
      ))
    })
  )
pvals_long$y.position <- max(pvals_long$y.position) + seq_len(nrow(pvals_long)) * 0.04
pvals_long$p <- round(pvals_long$p, 3)

# Add y-axis limits
y_limits <- c(0, max(pvals_long$y.position) + 0.1) # Adjust as needed

# Visualization - boxplot
p <- plotColData(
  tse, x = group_var, y = index,
  show_boxplot = TRUE, show_violin = FALSE
) +
  ylim(y_limits) +  # Set y-axis limits
  theme(text = element_text(size = 10)) +
  stat_pvalue_manual(pvals_long, label = "p")

print(p)
```

### 
