---
title: "Beta diversity"
author:
  - name: Muluh
---

## Community composition

In the report, a suffix of 1~ timepoint 1 and 2~ timepoint 2

```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"), fig.pos = "H")
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

### Phylum

```{r community-composition-phyla}
#| fig-width: 20
#| fig-height: 9
#| out.width: "100%"
#| fig.align: "center"

# Visualization - composition bar plot
plotAbundance(
  altExp(tse, "phylum"),
  assay.type = "relabundance",
  # color.by = "phylum_sub",
  color_by = "phylum_sub",  
  group = "phylum",
  order.row.by = "abund", 
  order.col.by = "Firmicutes") +
  #order.row.by = "abund") +  
theme(
          axis.text.y = element_text(face = "bold", size = 12),
          axis.title.x = element_text(face = "bold", size = 14),
          axis.title.y = element_text(face = "bold", size = 14),
          legend.text = element_text(face = "bold", size = 14),
          legend.title = element_text(face = "bold", size = 14)
        ) + scale_y_continuous(label=scales::percent) +
  labs(ylab = "Abundance(%)")

plots <- lapply(sort(unique(tse$diet)), function(diet_id) {
    tse_diet <- tse[, tse$diet == diet_id]
    
    # Inner lapply for time points
    time_plots <- lapply(unique(tse_diet$time), function(time_point) {
        tse_filtered <- tse_diet[, tse_diet$time == time_point]
        lab <- paste0(unique(tse_filtered$diet), " at time ", time_point)
        
        # Plot abundance for the filtered tse
        plotAbundance(
            altExp(tse_filtered, "phylum"),
            assay.type = "relabundance",
            color.by = "phylum_sub",
            group = "phylum",
            order.row.by = "abund",
            order.col.by = "Firmicutes"
        ) +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14)
            ) + scale_y_continuous(label=scales::percent) +
          labs(title = lab, ylab = "Abundance(%)")
    })
    return(time_plots)
})

# Flatten the list of plots
plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, ncol = 2)
bar_plot

```

### Genus

```{r community-composition-genus}
#| fig-width: 20
#| fig-height: 9
#| out.width: "100%"
#| fig.align: "center"
# Visualization - composition bar plot
top_taxa <- getTop(altExp(tse, "genus"), top = 10, assay.type = "relabundance")
genus_renamed <- lapply(rowData(altExp(tse, "genus"))$genus, function(x){
    if (x %in% top_taxa) {x} else {"Other"}})
rowData(altExp(tse, "genus"))$genus_sub <- as.character(genus_renamed)
altExp(tse, "genus") <- agglomerateByVariable(altExp(tse, "genus"), 
                                   by = "rows", 
                                   f = "genus_sub")

tse_g <- altExp(tse, "genus")
plotAbundance(
  tse_g,
  assay.type = "relabundance",
  color_by = "genus_sub",  
  group = "genus",
  order.row.by = "abund",
  order.col.by = "Collinsella") +
theme(
          axis.text.y = element_text(face = "bold", size = 12),
          axis.title.x = element_text(face = "bold", size = 14),
          axis.title.y = element_text(face = "bold", size = 14),
          legend.text = element_text(face = "bold", size = 14),
          legend.title = element_text(face = "bold", size = 14)
        ) + scale_y_continuous(label=scales::percent) +
  labs(ylab = "Abundance(%)")
          

plots <- lapply(sort(unique(tse_g$diet)), function(diet_id) {
    tse_diet <- tse_g[, tse_g$diet == diet_id]
    
    # Inner lapply for time points
    time_plots <- lapply(unique(tse_diet$time), function(time_point) {
        tse_filtered <- tse_diet[, tse_diet$time == time_point]
        lab <- paste0(unique(tse_filtered$diet), " at time ", time_point)
        
        # Plot abundance for the filtered tse
        plotAbundance(
            tse_filtered,
            assay.type = "relabundance",
            color.by = "genus_sub",
            group = "genus",
            order.row.by = "abund",
            order.col.by = "Collinsella"
          
        ) +  labs(title = lab, ylab = "Abundance(%)") +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14)
            ) + scale_y_continuous(label=scales::percent)
         
    })
    return(time_plots)
})

# Flatten the list of plots
plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, ncol = 2)
bar_plot
```

### Heatmap

```{r heatmap}
#| fig.width: 24
#| fig.height: 12
#| out.width: "100%"
#| fig.align: "center"

tse_heat <-  altExp(tse, "genus")
tse_heat <- transformAssay(tse_heat, assay.type = "relabundance", method = "clr", pseudocount = TRUE)
tse_heat <- transformAssay(
    tse_heat, assay.type = "clr", MARGIN = "rows", method = "standardize",
    name = "clr_z")
# Plot heatMap with sechm
plots <- lapply(sort(unique(tse_heat$diet)), function(diet_id) {
    # Filter tse_heat for the current meal group
    tse_heat_filtered <- tse_heat[, tse_heat$diet == diet_id]
    
    # Now iterate over unique times within the current meal group
    lapply(unique(tse_heat_filtered$time), function(time_point) {
        # Filter data for the current time
        tse_time_filtered <- tse_heat_filtered[, tse_heat_filtered$time == time_point]
        
        # Create the label based on both meal group and time
        lab <- paste0("Diet: ", unique(tse_time_filtered$diet), 
                      " at time: ", time_point)
        
        # Title grob
        title_grob <- textGrob(
            lab, x = 0.05, y = 0.2, just = c("left", "top"), 
            gp = gpar(fontsize = 14, fontface = "bold")
        )
        
        # Create heatmap for the filtered data
        heatmap <- sechm(
            tse_time_filtered, assayName = "clr_z", features = rownames(tse_time_filtered),
            show_rownames = TRUE, show_colnames = FALSE, do.scale = TRUE, name = "clr_z",
            top_annotation = character(), column_labels = tse_time_filtered$id,
            row_names_gp = gpar(fontsize = 12), column_names_gp = gpar(fontsize = 12),
            breaks = 1, column_title = "Samples", row_title = "Genus",
            width = unit(9, "inches"),
            height = unit(3, "inches"), hmcols = c("blue", "white", "red"), sort.method = "MDS_angle", cluster_cols = T, cluster_rows = T, na_col = "black"
        )
        
        # Grab heatmap as a grob
        heatmap_grob <- grid.grabExpr(draw(heatmap))
        
        # Combine title and heatmap
        combined_grob <- arrangeGrob(
            grobs = list(title_grob, heatmap_grob),
            ncol = 1,
            heights = unit.c(unit(1, "lines"), unit(1, "null"))
        )
        
        return(combined_grob)
    })
})


plots_flat <- unlist(plots, recursive = FALSE)

# Arrange all heatmaps in a grid
grid.newpage()
do.call(grid.arrange, c(plots_flat, ncol = 2))
```

## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# Define variables
tax.level <- "species"
dissimilarity.index <- "bray"
variable <- "diet"  
adjust.method <- "fdr"
test <- "Wilcoxon"
index <- "divergence" 
```

-   **Taxonomic level:** `r tax.level`
-   **Dissimilarity index:** `r dissimilarity.index`
-   **Group variable:** `r variable`

### PCoA plots

```{r pcoa-plot, fig.show="hold"}

tse <- runMDS(
        tse,
        FUN = getDissimilarity,
        method = "bray",
        assay.type = "relabundance",
        name = "PCoA_BC"
    )
tse <- transformAssay(tse, assay.type = "relabundance", method = "clr", pseudocount = TRUE)
# Calculate explained variance
e <- attr(reducedDim(tse, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])
# Create the PCoA plot
p <- plotReducedDim(tse, "PCoA_BC", 
                    colour_by = "meal", 
                    other_fields = c("diet", "group", "timepoint", "intervention", "meal", "time"))
# Add explained variance for axes
p1 <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
) + facet_wrap(diet ~ timepoint, nrow = 2)
# Add ellipses for groups
p_ellipse <- p1 + #stat_ellipse(aes(color = tse$group), level = 0.95) +
  theme(
      legend.position = "none",
      legend.box.margin = margin(-10, 0, 0, 0)
  )

p_ellipse
```

```{r trajectories}
# Plot 2: Trajectories
tse <- addStepwiseDivergence(tse, group = "id",
                            time.col = "time",
                            assay.type = "relabundance")
tse_sub <- tse[, !is.na(colData(tse)$divergence)]

# p2_traj <- plotReducedDim(
#     tse[, !is.na(colData(tse)[["meal_group"]])],
#     "PCoA_BC", colour_by = "timepoint",
#     other_fields = c("group", "id", "diet", "timepoint")
# ) +
#     aes(color = timepoint) +
#     labs(
#         title = "Trajectories between Treatments",
#         x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
#         y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
#     ) +
#     theme(plot.title = element_text(hjust = 0.5)) +
#     scale_color_manual(
#         values = c("before" = "blue", "after" = "orange"),
#         name = "Timepoint"
#     ) +
#     facet_wrap(~diet, nrow = 1) +
#     geom_line(
#         aes(group = id),
#         color = "grey50",
#         alpha = 0.5,
#         linetype = "dashed"
#     )
```

```{r run_diversity_tests}
# Calculate p-values
run_divergence_tests <- function(tse, index = "divergence") {
  # Extract values and timepoint
  values <- colData(tse)[[index]]
  timepoint <- tse$timepoint
  
  # Subset data for "after" timepoint only
  after_values <- values[timepoint == "after"]
  after_treatment <- tse$diet[timepoint == "after"]
  
  # Create an empty results dataframe
  results <- data.frame(
    Diet = character(),
    mean = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through unique treatment groups
  for (g_id in unique(after_treatment)) {
    # Subset the data for current treatment group
    subset_values <- after_values[after_treatment == g_id]
    
    # Calculate the mean for "after" timepoint
    mean_after <- mean(subset_values, na.rm = TRUE)
    
    # Perform Wilcoxon test (test if the median of "after" diverges from zero)
    test_result <- wilcox.test(
      subset_values,
      mu = 0,
      exact = FALSE,
      na.action = na.omit
    )
    
    # Add results to dataframe
    results <- rbind(results, data.frame(
      Diet = g_id,
      mean = mean_after,
      p.value = test_result$p.value,
      stringsAsFactors = FALSE
    )) %>%
      arrange(Diet)
  }
  
  return(results)
}

```

```{r pvalues-paired}
# Calculate statistical results
results_unpaired <- run_divergence_tests(tse_sub, index)

results_unpaired$p.adj <- p.adjust(results_unpaired$p.value, method = "fdr")
results_unpaired <- results_unpaired %>% arrange(desc(Diet))
```

```{r changes1}
div_change <- plotColData(
    tse_sub, x = "timepoint", y = index, colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("group", "diet", "timepoint")) + 
    facet_wrap(~diet, ncol = 4, scales="fixed", axis.labels = "margins") +
    scale_color_manual(
        values = c("after" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        legend.position = "none",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
    ) +
    labs(
        title = "Change",
        x = "",
        y = "Divergence over time"
    )
```

```{r viz-div_change}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
div_change
```

### Table of the top findings

##### Paired treatment between timepoint(1 and 2)

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **paired `r test`**

```{r table-of-findings-paired}
#| label: table-of-findings-paired
if (knitr::is_html_output()) {
  datatable(
    results_unpaired, 
    options = list(
      pageLength = 2,
      dom = 'Bfrtip'
    ),
    caption = "Table of divergence paired between timepoint(1 and 2)",
    rownames = FALSE
) %>% 
    formatRound(columns = c("mean"), digits = 2) %>%
  formatSignif(columns = c("p.adj", "p.value"), digits = 3)
} else {
  library(kableExtra)
  results_unpaired %>% 
    kbl(caption = "Table of divergence paired between timepoint(1 and 2)",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

```{r viz-fig2_traj}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
# p2_traj
```



```{r significant variables}
### Significant variables
# tse1 <- tse[, tse$diet == "rice"]
# tse2 <- tse[, tse$diet == "oat"]
# 
# # Perform RDA
# tse1 <- addRDA(
#     tse1,
#     assay.type = "relabundance",
#     formula = assay ~ timepoint ,
#     distance = dissimilarity.index,
#     na.action = na.exclude)
# 
# tse2 <- addRDA(
#     tse2,
#     assay.type = "relabundance",
#     formula = assay ~ timepoint,
#     distance = dissimilarity.index,
#     na.action = na.exclude)
```

```{r}
#| layout-ncol: 2
#| fig-width: 6
#| fig-height: 5
# tse_list <- list(tse1, tse2)
# Calculate explained variance
# lapply(tse_list, function(tse) {
# 
# e_rda1 <- attr(reducedDim(tse, "RDA"), "eig")
# rel_eig_rda1 <- e_rda1 / sum(e_rda1[e_rda1 > 0])
# # Generate RDA plot coloured by Treatment
# plotRDA(tse, "RDA") + 
#   labs(
#     x = paste("RDA 1 (", round(100 * rel_eig_rda1[[1]], 1), "%)", sep = ""),
#     y = paste("RDA 2 (", round(100 * rel_eig_rda1[[2]], 1), "%)", sep = "")
#   ) +
#   labs(color = "Diet") +
#   theme(legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10))
# 
# })

```



```{r mds}
#| fig-width: 10
#| fig-height: 3
#| width: 100%
#| fig.align: "center"
# -   MDS
# lapply(tse_list, function(tse) {
# pcoa_axes <- reducedDim(tse, "PCoA_BC") 
# taxa_abundances <- t(assay(tse, "clr")) 
# 
# # Calculate Kendall correlations
# cor_matrix <- t(cor(pcoa_axes, taxa_abundances, method = "kendall"))
# top_n <- 5
# top_taxa <- apply(cor_matrix, 2, function(x) {
#   abs_corr <- abs(x)
#   names(sort(abs_corr, decreasing = TRUE)[1:top_n])
# })
# 
# # Get unique taxa across all axes
# unique_taxa <- unique(unlist(top_taxa))
# filtered_cor_matrix <- cor_matrix[unique_taxa,]
# rownames(filtered_cor_matrix) <- rowData(tse[unique_taxa,])$genus
# plotLoadings(filtered_cor_matrix, ncomponents = 2) + 
# labs(
#     title = unique(tse$diet)
#   )
# })
```

## Supervised analysis


```{r significant-taxa-oat}
#| fig-width: 10
#| fig-height: 3
#| width: 100%
#| fig.align: "center"

### Significant taxa

# -   RDA
# lapply(tse_list, function(tse) {
# loadings_matrix <- attr(reducedDim(tse, "RDA"), "species")
# rownames(loadings_matrix) <- rowData(tse)$genus
# plotLoadings(loadings_matrix, ncomponents = 2, n = 10) +
#   labs(
#     title = unique(tse$diet)
#   )
# })
```

### Permanova analysis

-   Dissimilarity index: **`r dissimilarity.index`**

-   Group variable: **`r variable`**

**Interpretation**

-   Homogeneity \< 0.05: The assumption of homogeneity of variance is violated (i.e., the variances are significantly different across groups).

-   Homogeneity ≥ 0.05: The assumption of homogeneity of variance is not violated (i.e., the variances are similar across groups).

```{r run_rdba_tests}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova <- function(tse, variable, comparison, var) {
    tse_subset <- tse[, colData(tse)[, variable] %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    
    formula <- as.formula(paste("X ~", var))
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = formula, 
                          permutations = 999)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res)
    group_row <- perm_df[rownames(perm_df) == var, ]
    
    result <- data.frame(
        Group1 = comparison[1],
        Group2 = comparison[2],
        P_value = group_row$"permanova.Pr..F.",
        Homogeneity = signif(tail(perm_df$homogeneity.Pr..F., 1), 3)
        #P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests <- function(tse, comparisons, variable, var) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova(tse, variable, comp, var)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_before-after}
# For the before treatment comparisons
final_results_permanova <- run_rdba_tests(tse, comparisons_before, variable, var = variable)
final_results_permanova$p.adj <- p.adjust(final_results_permanova$P_value, method = "fdr")
# For the after treatment comparisons
final_results_permanova_after <- run_rdba_tests(tse, comparisons_after, variable, var = variable)
final_results_permanova_after$p.adj <- p.adjust(final_results_permanova_after$P_value, method = "fdr")

final_results_permanova_tp <- run_rdba_tests(tse, comparisons_before, variable, var = "time")
final_results_permanova_tp$p.adj <- p.adjust(final_results_permanova_tp$P_value, method = "fdr")
# For the after treatment comparisons
final_results_permanova_after_tp <- run_rdba_tests(tse, comparisons_after, variable, var = "time")
final_results_permanova_after_tp$p.adj <- p.adjust(final_results_permanova_after_tp$P_value, method = "fdr")

#final_results_permanova[] <- lapply(final_results_permanova, function(x) gsub(".before|.after", "", x))
#final_results_permanova_after[] <- lapply(final_results_permanova_after, function(x) gsub(".before|.after", "", x))
```

```{r permanova_paired}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova_pair <- function(tse, variable, comparison, var) {
    tse_subset <- tse[, colData(tse)[, variable] %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    
    formula <- as.formula(paste("X ~", var))
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = formula, 
                          permutations = 999)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res)
    group_row <- perm_df[rownames(perm_df) == var, ]
    
    result <- data.frame(
        Diet = comparison,
        P_value = group_row$"permanova.Pr..F.",
        Homogeneity = signif(tail(perm_df$homogeneity.Pr..F., 1), 3)
        #P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests_pair <- function(tse, comparisons, variable, var) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova_pair(tse, variable, comp, var)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_paired_res}
# For the paired treatment comparisons
diet_groups <- sort(unique(tse$diet))
# Calculate statistical results
  
# Run the diversity tests
 # for (comp in meal_groups) {
 #    respair <- run_rdba_tests_pair(tse, comp, variable, var= "time")
 # }

final_results_permanova_paired<- lapply(diet_groups, function(diet_id) {
    respair <- run_rdba_tests_pair(tse, comp = diet_id, variable = "diet", var = "time")
    return(respair)
}) %>% 
  bind_rows() %>%
  arrange(Diet)
final_results_permanova_paired$p.adj <- p.adjust(final_results_permanova_paired$P_value, method = "fdr")

final_results_permanova_paired <- final_results_permanova_paired %>% arrange(desc(Diet))
```

#### Across-group

```{r permanova_res_before}
#| label: permanova_res_before

if (knitr::is_html_output()) {
  datatable(final_results_permanova,
            options = list(pageLength = 1),
            caption = "PERMANOVA Results - before treatment - X ~ diet")
} else {
  library(kableExtra)
  final_results_permanova %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - before treatment - X ~ diet",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

```{r permanova_res_beforetp}
#| label: permanova_res_beforetp

if (knitr::is_html_output()) {
  datatable(final_results_permanova_tp,
            options = list(pageLength = 1),
            caption = "PERMANOVA Results - before treatment - X ~ time")
} else {
  library(kableExtra)
  final_results_permanova_tp %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - before treatment - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_tp)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

```{r permanova_res_after}
#| label: permanova_res_after

if (knitr::is_html_output()) {
  datatable(final_results_permanova_after,
            options = list(pageLength = 1),
            caption = "PERMANOVA Results - after treatment - X ~ diet")
} else {
  library(kableExtra)
  final_results_permanova_after %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - after treatment - X ~ diet",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_after)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

```{r permanova_res_aftertp}
#| label: permanova_res_aftertp

if (knitr::is_html_output()) {
  datatable(final_results_permanova_after_tp,
            options = list(pageLength = 6),
            caption = "PERMANOVA Results - after treatment - X ~ time")
} else {
  library(kableExtra)
  final_results_permanova_after_tp %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - after treatment - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_after_tp)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

#### Within-group

```{r permanova_paired_tab}
#| label: permanova_paired_tab
if (knitr::is_html_output()) {
  datatable(final_results_permanova_paired,
            options = list(pageLength = 4),
            caption = "PERMANOVA Results - Paired after intervention - X ~ time")
} else {
  library(kableExtra)
  final_results_permanova_paired %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - Paired after intervention - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_paired)-1))) %>%
    row_spec(0, bold = TRUE)
}
 
   
```

```{r}
tse <- readRDS("../../data/tse.Rds")
tabs_species <- NULL 
tseg <- altExp(tse, "species_prevalent")
for (g in unique(tseg$diet)) {
    #tseg <- tseg[, tseg$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$species)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars

      tabs_species <-
        rbind(tabs_species,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          iqr95_1 = quantile(subset(df, timepoint == "before")$signal, 0.95),
          iqr95_2 = quantile(subset(df, timepoint == "after")$signal, 0.95),
	  logFC.rel = mean(subset(df, timepoint=="after")$signal - subset(df, timepoint=="before")$signal)
        ))
    }
}
tabs_species <- as.data.frame(tabs_species)
tabs_species$mean.rel1 <- round(100 * as.numeric(tabs_species$mean1), 3); tabs_species$mean1 <- NULL
tabs_species$mean.rel2 <- round(100 * as.numeric(tabs_species$mean2), 3); tabs_species$mean2 <- NULL
tabs_species$iqr95_1 <- round(100 * as.numeric(tabs_species$iqr95_1), 3); tabs_species$`iqr95_1.95%` <- NULL
tabs_species$iqr95_2 <- round(100 * as.numeric(tabs_species$iqr95_2), 3); tabs_species$`iqr95_2.95%` <- NULL
tabs_species$logFC.rel <- round(100 * as.numeric(tabs_species$logFC.rel), 3); 
```


```{r}
tabs_genus <- NULL 
 tseg <- altExp(tse, "genus_prevalent")
for (g in unique(tseg$diet)) {
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$genus)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars

      tabs_genus <-
        rbind(tabs_genus,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          iqr95_1 = quantile(subset(df, timepoint == "before")$signal, 0.95),
          iqr95_2 = quantile(subset(df, timepoint == "after")$signal, 0.95),
	  logFC.rel = mean(subset(df, timepoint=="after")$signal - subset(df, timepoint=="before")$signal)
        ))
    }
}
tabs_genus <- as.data.frame(tabs_genus)
tabs_genus$mean.rel1 <- round(100 * as.numeric(tabs_genus$mean1), 3); tabs_genus$mean1 <- NULL
tabs_genus$mean.rel2 <- round(100 * as.numeric(tabs_genus$mean2), 3); tabs_genus$mean2 <- NULL
tabs_genus$iqr95_1 <- round(100 * as.numeric(tabs_genus$iqr95_1), 3); tabs_genus$`iqr95_1.95%` <- NULL
tabs_genus$iqr95_2 <- round(100 * as.numeric(tabs_genus$iqr95_2), 3); tabs_genus$`iqr95_2.95%` <- NULL
tabs_genus$logFC.rel <- round(100 * as.numeric(tabs_genus$logFC.rel), 3); 
```

```{r}
#| label: tablespecies
if (knitr::is_html_output()) {
  datatable(tabs_species,
            options = list(pageLength = 10),
            caption = "Table of average relative abundance and 95th percentile - Species")
} else {
  kable(tabs_species)
}
```

```{r}
#| label: tablegenus
if (knitr::is_html_output()) {
  datatable(tabs_genus,
            options = list(pageLength = 10),
            caption = "Table of average relative abundance and 95th percentile - Genus")
} else {
  kable(tabs_genus)
}
```


### Data Summary

```{r data-summary}
tse
```
