---
title: "Bacteroidetes/Firmicutes ratio"
author:
  - name: Muluh
---

## Bacteroidetes/Firmicutes ratio

```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"))
# devtools::load_all("~/Rpackages/microbiome/miaverse/miaViz")
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

```{r setup, include=FALSE}
# Define variables
tax.level     <- "Phylum"
variable      <- "diet"  
adjust.method <- "fdr"
test          <- "Wilcoxon"
index         <- "bact_firm_ratio"
```

### Group-wise comparisons

-   **Taxonomic level:** `r tax.level`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

```{r bact-firm}
# Calculate the Bacteroidetes/Firmicutes ratio and add to colData
# using the chosen name
tse <- altExp(tse, "phylum")
ratio <- "bact_firm_ratio"
colData(tse)[[ratio]] <- assay(tse, "relabundance")["Bacteroidetes", ] /
                         assay(tse, "relabundance")["Firmicutes", ]
```

```{r run_diversity_tests}
# Calculate p-values
run_diversity_tests <- function(tse, comparisons, 
                                variable, ratio, 
                                adjust.method = "fdr", paired = FALSE) {
  # Get all values and groups
  values <- colData(tse)[[ratio]]
  groups <- colData(tse)[[variable]]
  
  # Create an empty results dataframe
  results <- data.frame(
    group1 = character(),
    group2 = character(),
    mean1 = numeric(),
    mean2 = numeric(),
    logFC = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through requested comparisons
  for (comp in comparisons) {
    group1 <- comp[1]
    group2 <- comp[2]
    
    # Subset data for just these two groups
    subset_indices <- which(groups %in% c(group1, group2))
    subset_values <- values[subset_indices]
    subset_groups <- groups[subset_indices]
    
    # Ensure paired data is properly aligned (if paired is TRUE)
    if (paired) {
      # Assuming 'subject' column exists for pairing
      subset_subjects <- tse$id[subset_indices]
      paired_data <- data.frame(
        subject = subset_subjects,
        group = subset_groups,
        value = subset_values
      ) %>%
        tidyr::pivot_wider(names_from = group, values_from = value) %>%
        drop_na()  # Drop pairs with missing data
      
      # Extract paired values
      paired_values <- paired_data %>% select(all_of(c(group1, group2)))
    }
    
    # Perform Wilcoxon test
    if (paired) {
      test_result <- wilcox.test(
        paired_values[[1]], paired_values[[2]],
        paired = TRUE, exact = FALSE
      )
    } else {
      test_result <- wilcox.test(
        subset_values ~ subset_groups,
        exact = FALSE
      )
    }
    
    # Calculate means
    mean1 <- mean(values[groups == group1])
    mean2 <- mean(values[groups == group2])

    if (paired) {
       # Note: when the values are paired, we can calculate FC separately
       # for every individual and summarize afterwards. 
       FC <- mean(values[groups == group2] / values[groups == group1])
    } else {
       FC <- mean2/mean1
    }

    # Add results
    results <- rbind(results, data.frame(
      group1 = group1,
      group2 = group2,
       mean1 = mean1,
       mean2 = mean2,
       logFC = log2(FC),
      p.value = test_result$p.value,
      stringsAsFactors = FALSE
    ))

  }
  
  # Add adjusted p-values using specified method
  # results$p.adj <- p.adjust(results$p.value, method = adjust.method)
  
  return(results)
}

```

```{r pvalues-before}
tse_before <- tse[, tse$visit == 1]
# Calculate statistical results
results_before <- run_diversity_tests(tse_before, comparisons_before, 
                                      variable, ratio = "bact_firm_ratio", 
                                      adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 2)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse_before)[[ratio]][colData(tse_before)[[variable]] == group1[i]]
    vals2 <- colData(tse_before)[[ratio]][colData(tse_before)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
tse_after <- tse[, tse$visit == 2]
# Calculate statistical results
results_after <- run_diversity_tests(tse_after, comparisons_after, 
                                     variable, ratio = "bact_firm_ratio", 
                                     adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 2)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse_after)[[ratio]][colData(tse_after)[[variable]] == group1[i]]
    vals2 <- colData(tse_after)[[ratio]][colData(tse_after)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r pvalues-paired}
diet_groups <- unique(tse$diet)
# Calculate statistical results
results_paired <- lapply(diet_groups, function(group) {
  # Subset tse for the current meal group
  tse_subset <- tse[, tse$diet == group]
  
  # Run the diversity tests
  res <- run_diversity_tests(
    tse = tse_subset, 
    comparisons = comparisons_paired, 
    variable = "timepoint", 
    ratio = "bact_firm_ratio", 
    adjust.method = adjust.method, 
    paired = TRUE
  )
  
  # Add the meal group as a new column
  res$diet <- group
  
  return(res)
}) %>% 
  bind_rows()

results_paired <- results_paired[, c("diet", setdiff(names(results_paired), "diet"))]

# results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.value = round(p.value, 2)
  ) %>%
  select(diet, group1, group2, mean1, mean2, logFC, p.value) %>% 
  arrange(diet)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[ratio]][colData(tse)[["timepoint"]] == group1[i]]
    vals2 <- colData(tse)[[ratio]][colData(tse)[["timepoint"]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```

```{r bact-firm-before}
p1 <- plotColData(
    tse_before, x = variable, y = "bact_firm_ratio", colour_by = "timepoint",
    show_boxplot = TRUE, other_fields = c("timepoint"),
    show_violin = FALSE) +
    scale_color_manual(
        values = c("before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # Remove legend
        legend.position = "none"
    ) +
    ggpubr::stat_pvalue_manual(
      data = final_results_before,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "Before treatment",
      x = "",
      y = "Bacteroidetes/Firmicutes ratio"
     ) #+ scale_y_log10()
```

```{r bact-firm-after}
p2 <- plotColData(
    tse_after, x = variable, y = "bact_firm_ratio", colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint")) +
    scale_color_manual(
        values = c("after" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) +
    ggpubr::stat_pvalue_manual(
      data = final_results_after,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "After treatment",
      x = "",
      y = "Bacteroidetes/Firmicutes ratio"
    ) #+ scale_y_log10()
```

```{r all-groups-1v2}
p3 <- plotColData(
    tse, x = "timepoint", y = "bact_firm_ratio", colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint", "diet")) +
    scale_color_manual(
        values = c("after" = "black", "before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 2)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )  +
    labs(
      title = "Within group",
      x = "",
      y = "Bacteroidetes/Firmicutes ratio"
    ) +
    facet_wrap(~diet, ncol = 4, scales = "free")

```

```{r bact-firm-change}
tse_new <- tse[,tse$paired == "yes"]
colData(tse_new) <- DataFrame(
    colData(tse_new) %>%
        as_tibble() %>%
        tidyr::drop_na({{ variable }}) %>%
        filter(paired == "yes") %>%
        group_by(id) %>%
        mutate(
            change = if_else(time == 2, .data[[index]] - .data[[index]][time == 1], 0),
            change = if_else(time == 1, change[time == 2], change)
        ) %>%
        ungroup() %>%
        as.data.frame(check.names = FALSE)
)

div_change <- plotColData(
    tse_new, x = "timepoint", y = "bact_firm_ratio",
    show_boxplot = TRUE,
    show_violin = FALSE,
    other_fields = c("group", "id", "diet", "change")) + 
    facet_wrap(~diet, ncol = 4, scales="fixed", axis.labels = "margins") +
    theme(
        text = element_text(size = 10),
    ) +
    labs(
        title = "Paired Bacteroidetes/Firmicutes ratio",
        x = "",
        y = "Bacteroidetes/Firmicutes ratio"
    ) + 
    geom_line(
        aes(group = id, color = change)
    ) +
    scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab") +
    #scale_y_log10() +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 2)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )
```

```{r visualize-plot}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
y_min <- min(c(tse_before$bact_firm_ratio, tse_after$bact_firm_ratio), na.rm = TRUE)
y_max <- max(c(tse_before$bact_firm_ratio, tse_after$bact_firm_ratio), na.rm = TRUE)

# Get the maximum y-position from your annotations
max_annotation_y <- max(c(
    final_results_before$y.position, 
    final_results_after$y.position
), na.rm = TRUE)

# Use the larger of data max and annotation position for the upper limit
y_max <- max(y_max, max_annotation_y)

y_range <- y_max - y_min
y_max <- y_max + (y_range * 0.13)  

# Apply to both plots
p1 <- p1 + coord_cartesian(ylim = c(y_min, y_max))
p2 <- p2 + coord_cartesian(ylim = c(y_min, y_max))
#wrap_plots(p1, p2, axes = "collect")
```

#### Table of the top findings

#### Across-group

##### Before treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-before}
#| label: table-of-findings-before

if (knitr::is_html_output()) {
  datatable(
      final_results_before %>% select(-y.position), 
      options = list(
        pageLength = 6,
        dom = 'Bfrtip'
      ),
      caption = "Table of  Bacteroidetes/Firmicutes before Treatment",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_before %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidetes/Firmicutes before Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

##### After treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-after}
#| label: table-of-findings-after
if (knitr::is_html_output()) {
  datatable(
    final_results_after %>% select(-y.position), 
    options = list(
      pageLength = 6,
      dom = 'Bfrtip'
    ),
    caption = "Table of Bacteroidetes/Firmicutes after Treatment",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_after %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidetes/Firmicutes after Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}

```

```{r viz-div_change}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
div_change
```

#### Within-group

##### Paired treatment

-   Multiple testing correction: **paired** -none
-   Statistical test: **`r test`**

```{r table-of-findings-paired}
#| label: table-of-findings-paired
if (knitr::is_html_output()) {
  datatable(
    final_results_paired %>% select(-y.position), 
    options = list(
      pageLength = 4,
      dom = 'Bfrtip'
    ),
    caption = "Table of Bacteroidetes/Firmicutes paired between timepoint (1 and 2)",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.value", digits = 3)
} else {
  library(kableExtra)
  final_results_paired %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidetes/Firmicutes paired between timepoint (1 and 2)",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

```{r lmer-stuff}
tse$bfratio <- log2(tse$bact_firm_ratio)
res <- run_lmer(tse, "bfratio")
l <- summary(res)
```

```{r clean-lmer}
fixed_effects <- as.data.frame(l$coefficients) %>%
  rownames_to_column("Term") 
```

```{r table-fixed-fx}
#| label: table-fx
if (knitr::is_html_output()) {
  datatable(
    fixed_effects, 
    options = list(
      pageLength = 4,
      dom = 'Bfrtip'
    ),
    caption = "Table of interaction effects",
    rownames = FALSE
)
} else {
  library(kableExtra)
  fixed_effects %>% 
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of interaction effects",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

## Data Summary

```{r data-summary}
tse
```

```{r sessioninfo}
session_info <- capture.output(sessionInfo())
output_path <- "../../output/sesinfo.txt"
writeLines(session_info, output_path)
```
