---
title: "Alpha Diversity Analysis: Observed Species"
author: "Nitin Bayal"
format: html
editor: visual
execute:
  echo: true
---

## Introduction

This document reports on the results of alpha diversity analyses, focusing on observed species across different diets and visits.

## Setup

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r data-load}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(TreeSummarizedExperiment)
library(mia)

# Source custom functions from provided scripts
source("main.R")
source("primary2Groups_funct.R")
source("TreeSE.R")

# Load dataset
diversity_data <- read_csv("./data/merged_diversity_results.csv")

# Load TreeSummarizedExperiment (TSE) object if it was saved separately
tse <- readRDS("./data/tse.Rds")
```

## Data Summary

```{r data-summary}
#| echo: false
# Summarize the loaded data
summary(diversity_data)
```

The `echo: false` option disables the printing of code (only output is displayed).

## Alpha Diversity: Observed Species

### Observed Species Analysis Between Groups 

#### Diet 1, Visit 1 vs. Diet 1, Visit 2 

The observed species comparison between diet 1, visit 1 and diet 1, visit 2 cab be visualized below:

```{r observed-d1v1-d1v2}
# Create the plot for Diet 1, Visit 1 vs Diet 1, Visit 2
create_diversity_plot(tse, c("diet_1_visit_1", "diet_1_visit_2"), "observed", "output/")
```

#### Diet 1, Visit 1 vs. Diet 2, Visit 1 

The observed species comparison between diet 1, visit 1 and diet 2, visit 1 can be visualized below:

```{r observed-d1v1-d2v1}
# Create the plot for Diet 1, Visit 1 vs Diet 2, Visit 1
create_diversity_plot(tse, c("diet_1_visit_1", "diet_2_visit_1"), "observed", "output/")
```

#### Diet 1, Visit 2 vs. Diet 2, Visit 2

The observed species comparison between diet 1, visit 2 and diet 2, visit 2 can be visualized below:

```{r observed-d1v2-d2v2}
# Create the plot for Diet 1, Visit 2 vs Diet 2, Visit 2
create_diversity_plot(tse, c("diet_1_visit_2", "diet_2_visit_2"), "observed", "output/")
```

#### Diet 2, Visit 1 vs. Diet 2, Visit 2 

The observed species comparison between diet 2, visit 1 and diet 2, visit 2 can be visualized below:

```{r observed-d2v1-d2v2}
# Create the plot for Diet 1, Visit 1 vs Diet 2, Visit 1
create_diversity_plot(tse, c("diet_2_visit_1", "diet_2_visit_2"), "observed", "output/")
```

### Statistical Analysis

To determine if the observed diversity differences between the groups are statistically significant, a Wilcoxon test is performed.

```{r statistical-analysis}
# Perform a Wilcoxon test for observed species between diet 1, visit 1 and diet 1, visit 2
test_result <- perform_significance_test(tse, c("diet_1_visit_1", "diet_1_visit_2"), "group", "observed")
print(test_result)

# Perform a Wilcoxon test for observed species between diet 1, visit 1 and diet 2, visit 1
test_result <- perform_significance_test(tse, c("diet_1_visit_1", "diet_2_visit_1"), "group", "observed")
print(test_result)

# Perform a Wilcoxon test for observed species between diet 1, visit 2 and diet 2, visit 2
test_result <- perform_significance_test(tse, c("diet_1_visit_2", "diet_2_visit_2"), "group", "observed")
print(test_result)

# Perform a Wilcoxon test for observed species between diet 2, visit 1 and diet 2, visit 2
test_result <- perform_significance_test(tse, c("diet_1_visit_1", "diet_1_visit_2"), "group", "observed")
print(test_result)
```

### Results and Discussion

The alpha diversity analysis using observed species indicates variability across diets and visits. Significant differences highlight the impact of dietary interventions over time.
