---
title: "Alpha Diversity"
author:
  - name: Muluh
format:
  html:
    code-fold: true
    number-sections: true
    page-layout: full
editor: visual
params: 
  index: default
execute:
  echo: false
  warning: false
  message: false
---

## Community diversity

```{r variables}
# Alpha diversity
index <- params$index
variable      <- "diet" 
adjust.method <- "fdr"
test <- "Wilcoxon"
```

```{r data}
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

## Group-wise comparisons

-   **Index:** `r params$index`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

```{r run_diversity_tests}
# Calculate p-values
run_diversity_tests <- function(tse, comparisons, 
                                variable, index, 
                                adjust.method = "fdr", paired = FALSE) {
  # Get all values and groups
  values <- colData(tse)[[index]]
  groups <- colData(tse)[[variable]]
  
  # Create an empty results dataframe
  results <- data.frame(
    group1 = character(),
    group2 = character(),
    mean1 = numeric(),
    mean2 = numeric(),
    logFC = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through requested comparisons
  for (comp in comparisons) {
    group1 <- comp[1]
    group2 <- comp[2]
    
    # Subset data for just these two groups
    subset_indices <- which(groups %in% c(group1, group2))
    subset_values <- values[subset_indices]
    subset_groups <- groups[subset_indices]
    
    # Ensure paired data is properly aligned (if paired is TRUE)
    if (paired) {
      # Assuming 'subject' column exists for pairing
      subset_subjects <- tse$id[subset_indices]
      paired_data <- data.frame(
        subject = subset_subjects,
        group = subset_groups,
        value = subset_values
      ) %>%
        tidyr::pivot_wider(names_from = group, values_from = value) %>%
        drop_na()  # Drop pairs with missing data
      
      # Extract paired values
      paired_values <- paired_data %>% select(all_of(c(group1, group2)))
    }
    
    # Perform Wilcoxon test
    if (paired) {
      test_result <- wilcox.test(
        paired_values[[1]], paired_values[[2]],
        paired = TRUE, exact = FALSE
      )
    } else {
      test_result <- wilcox.test(
        subset_values ~ subset_groups,
        exact = FALSE
      )
    }
    
    # Calculate means
    mean1 <- mean(values[groups == group1])
    mean2 <- mean(values[groups == group2])

    if (paired) {
       # Note: when the values are paired, we can calculate FC separately
       # for every individual and summarize afterwards. 
       FC <- mean(values[groups == group2] / values[groups == group1])
    } else {
       FC <- mean2/mean1
    }

    # Add results
    results <- rbind(results, data.frame(
      group1 = group1,
      group2 = group2,
       mean1 = mean1,
       mean2 = mean2,
       logFC = log2(FC),
      p.value = test_result$p.value,
      stringsAsFactors = FALSE
    ))

  }
  
  # Add adjusted p-values using specified method
  # results$p.adj <- p.adjust(results$p.value, method = adjust.method)
  
  return(results)
}
```

```{r pvalues-before}
tse_before <- tse[, tse$visit == 1]
# Calculate statistical results
results_before <- run_diversity_tests(tse_before, comparisons_before, variable, index, adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse_before)[[index]][colData(tse_before)[[variable]] == group1[i]]
    vals2 <- colData(tse_before)[[index]][colData(tse_before)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
tse_after <- tse[, tse$visit == 2]
# Calculate statistical results
results_after <- run_diversity_tests(tse_after, comparisons_after, variable, index, adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse_after)[[index]][colData(tse_after)[[variable]] == group1[i]]
    vals2 <- colData(tse_after)[[index]][colData(tse_after)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r pvalues-paired}
diet_groups <- unique(tse$diet)
# Calculate statistical results
results_paired <- lapply(diet_groups, function(group) {
  # Subset tse for the current meal group
  tse_subset <- tse[, tse$diet == group]
  
  # Run the diversity tests
  res <- run_diversity_tests(
    tse = tse_subset, 
    comparisons = comparisons_paired, 
    variable = "timepoint", 
    index, 
    adjust.method = adjust.method, 
    paired = TRUE
  )
  
  # Add the meal group as a new column
  res$diet <- group
  
  return(res)
}) %>% 
  bind_rows()

results_paired <- results_paired[, c("diet", setdiff(names(results_paired), "diet"))]

# results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.value = round(p.value, 3)
  ) %>%
  select(diet, group1, group2, mean1, mean2, logFC, p.value) %>% 
  arrange(diet)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[index]][colData(tse)[["timepoint"]] == group1[i]]
    vals2 <- colData(tse)[[index]][colData(tse)[["timepoint"]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```

```{r boxplot-before}
# Create a boxplot
p1 <- plotColData(
    tse_before, x = variable, y = index, colour_by = "timepoint",
    show_boxplot = TRUE, other_fields = c("timepoint"),
    show_violin = FALSE) +
    scale_color_manual(
        values = c("before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # Remove legend
        legend.position = "none"
    ) + coord_fixed(ratio = 0.5) +
    ggpubr::stat_pvalue_manual(
      data = final_results_before,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "Before treatment",
      x = "",
      y = paste("Alpha Diversity -", index)
     ) #+ scale_y_log10()
```

```{r boxplot-after}
# Create a boxplot
p2 <- plotColData(
    tse_after, x = variable, y = index, colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint")) + 
    scale_color_manual(
        values = c("after" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) + coord_fixed(ratio = 0.5) +
    ggpubr::stat_pvalue_manual(
      data = final_results_after,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "After treatment",
      x = "",
      y = paste("Alpha Diversity -", index)
    ) #+ scale_y_log10()
```

```{r all-groups-1v2}
p3 <- plotColData(
    tse, x = "timepoint", y = index, colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint", "diet")) +
    scale_color_manual(
        values = c("after" = "black", "before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 3)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )  +
    labs(
      title = "Within group",
      x = "",
      y = paste("Alpha Diversity -", index)
    ) +
    facet_wrap(~diet, ncol = 4, scales = "free")
```

```{r changes1}
tse_new <- tse[,tse$paired == "yes"]
colData(tse_new) <- DataFrame(
    colData(tse_new) %>%
        as_tibble() %>%
        tidyr::drop_na({{ variable }}) %>%
        filter(paired == "yes") %>%
        group_by(id) %>%
        mutate(
            change = if_else(time == 2, .data[[index]] - .data[[index]][time == 1], 0),
            change = if_else(time == 1, change[time == 2], change)
        ) %>%
        ungroup() %>%
        as.data.frame(check.names = FALSE)
)

div_change <- plotColData(
    tse_new, x = "timepoint", y = index,
    show_boxplot = TRUE,
    show_violin = FALSE,
    other_fields = c("group", "id", "diet", "change")) + 
    facet_wrap(~diet, ncol = 4, scales="fixed", axis.labels = "margins") +
    theme(
        text = element_text(size = 10),
    ) +
    labs(
        title = "Paired alpha diversity",
        x = "",
        y = paste("Alpha Diversity -", index)
    ) + 
    geom_line(
        aes(group = id, color = change)
    ) +
    scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab") +
    #scale_y_log10() +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 3)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )

```

```{r visualize-plot}
#| fig-width: 5
#| fig-height: 4
#| fig.align: "center"
y_min <- min(c(tse_before[[index]], tse_after[[index]]), na.rm = TRUE)
y_max <- max(c(tse_before[[index]], tse_after[[index]]), na.rm = TRUE)

# Get the maximum y-position from your annotations
max_annotation_y <- max(c(
    final_results_before$y.position, 
    final_results_after$y.position
), na.rm = TRUE)

# Use the larger of data max and annotation position for the upper limit
y_max <- max(y_max, max_annotation_y)

y_range <- y_max - y_min
y_max <- y_max + (y_range * 0.2)  

# Apply to both plots
p1 <- p1 + coord_cartesian(ylim = c(y_min, y_max))
p2 <- p2 + coord_cartesian(ylim = c(y_min, y_max))
wrap_plots(p1, p2, axes = "collect")
```

### Table of the top findings

#### Across-group

##### Before treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-before}
#| label: table-of-findings-before
if (knitr::is_html_output()) {
  datatable(
      final_results_before %>% select(-y.position), 
      options = list(
        pageLength = 6,
        dom = 'Bfrtip'
      ),
      caption = "Table of alpha diversity comparisons before Treatment",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_before %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity comparisons before Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

##### After treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-after}
#| label: table-of-findings-after
if (knitr::is_html_output()) {
  datatable(
    final_results_after %>% select(-y.position), 
    options = list(
      pageLength = 6,
      dom = 'Bfrtip'
    ),
    caption = "Table of alpha diversity comparisons after Treatment",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_after %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity comparisons after Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}

```

```{r viz-div_change}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
div_change
```

#### Within-group

##### Paired treatment

-   Multiple testing correction: **paired** -none
-   Statistical test: **`r test`**

```{r table-of-findings-paired}
#| label: table-of-findings-paired
if (knitr::is_html_output()) {
  datatable(
    final_results_paired %>% select(-y.position), 
    options = list(
      pageLength = 4,
      dom = 'Bfrtip'
    ),
    caption = "Table of alpha diversity paired between timepoint(1 and 2)",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.value", digits = 3)
} else {
  library(kableExtra)
  final_results_paired %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity paired between timepoint(1 and 2)",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

## Alternate analyses

```{r stratified-timepoint}
# First analysis - stratified by time point
tbl_a_s <- colData(tse) %>% as_tibble() %>%
    select(c("timepoint", index, "meal_group", "age")) %>%
    tbl_strata(strata = timepoint, .combine_with = "tbl_merge",
               .tbl_fun = ~.x %>% tidyr::drop_na() %>% 
                   lm(formula = as.formula(
                       paste0(index, " ~ meal_group + age")),
                      data = .) %>% 
                   tbl_regression(
                       intercept = TRUE, 
                       include = everything()) %>%
                   add_global_p(quite=TRUE, keep=TRUE))
```

```{r paired}
# Paired analysis
tbl_a_s_diff <- colData(tse) %>% as_tibble() %>%
    select(c("timepoint", index, variable, "meal_group", "paired", "age")) %>% 
    tidyr::drop_na(variable) %>% filter(paired=="yes") %>% 
    lmerTest::lmer(
        formula = as.formula(
              paste0(index, " ~ meal_group + age + (1|", variable, ")")
        ),
        data = .) %>% 
    tbl_regression(
        intercept = TRUE, 
        include = everything()) %>%
    add_global_p(quite=TRUE, keep=TRUE) %>% 
    modify_header(label = "**Paired test between timepoints (1 and 2)**")
```

-   **Beta (Estimate)**: Shows the effect size and direction of a predictor on the outcome. Positive means an increase, negative a decrease.

-   **95% CI**: Range where the true beta is likely to fall, reflecting precision.

-   **p-Value**: Tests if beta differs from zero; p\<0.05p \< 0.05p\<0.05 suggests significance.

```{r table-res1}
knitr::knit_print(tbl_a_s)
```

```{r table-res2}
knitr::knit_print(tbl_a_s_diff)
```

```{r Plotting-code}
# Plotting code
# Set up reference levels
ref <- levels(factor(colData(tse)[["meal_group"]]))[1]
lvs <- levels(factor(colData(tse)[["meal_group"]]))[-1]
comps <- lapply(lvs, function(lv) c(ref,lv))
```

```{r first-plot}
# First plot
p1 <- colData(tse) %>% as_tibble() %>%  tidyr::drop_na(variable) %>%
    as.data.frame(check.names=FALSE) %>%
    ggplot(aes_string(x="meal_group", y = index))+
    geom_boxplot(outlier.shape = NA)+
    geom_jitter(width=0.1, size=0.5)+
    facet_wrap(~timepoint, ncol = 2, scales="free")+
    geom_signif(comparisons = comps, map_signif_level = FALSE,
                step_increase=0.2, textsize = 2.5)+
    labs(x="", y = paste(index, "index"))

# Process p-values for plot
p_tmp <- ggplot_build(p1)
df_1 <- p_tmp$data[[3]] %>% mutate(label=gsub(paste0(ref,"|\\-|[0-9]"),"",
                                              .data[["group"]]))
df_2 <- tbl_a_s$table_body %>% filter(label %in% lvs) %>%
    select(c(label, contains("p.value"))) %>% 
    tidyr::pivot_longer(cols = contains("p.value"),
                        names_to="PANEL", names_prefix="p.value_") %>% 
    mutate(PANEL=factor(PANEL, levels = levels(df_1$PANEL)))
p_tmp$data[[3]] <- merge(df_1,df_2, by = c("PANEL","label")) %>%
    mutate(annotation=round(value,3)) %>% select(colnames(p_tmp$data[[3]]))
p1 <- p_tmp %>% ggplot_gtable() %>% ggplotify::as.ggplot()


```

```{r second-plot}
# Second plot
df_pval <- tbl_a_s_diff$table_body %>% filter(label %in% lvs) %>%
    select(c(label, contains("p.value")))
pos_j <- position_jitter(width = 0.1, height = 0)
p2 <- colData(tse) %>% as_tibble() %>%  tidyr::drop_na({{ variable }}) %>%
    filter(paired=="yes") %>% group_by(.data[[variable]]) %>% arrange(id, timepoint) %>%
    mutate(
        change = .data[[index]] - dplyr::first(.data[[index]])
    ) %>% tidyr::fill(change, .direction = "downup") %>%  
    as.data.frame(check.names=FALSE) %>%
    ggplot(aes(x = timepoint, y = .data[[index]]))+
    geom_boxplot(outlier.shape = NA)+
    geom_point(position = pos_j, size = 0.5) + 
    geom_line(aes(group = id, color = change), position = pos_j) +
    scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab")+
    facet_wrap(as.formula(paste0("~", "meal_group")), nrow = 1,
               labeller = function(df) {
                   df[,1] <- sapply(df[,1], function(lb){
                       if (lb!=ref) {
                           lb <- paste0(lb, " (p-val = ",
                                        df_pval %>% filter(label==lb) %>%
                                            select(contains("p.value")) %>%
                                            .[[1]] %>% round(3)
                                        ,")")
                           lb
                       } else {
                           as.character(lb)
                       }
                   }) %>% unlist()
                   df
               })+
    labs(x="", y = paste(index, "index"))


```

## Data Summary

```{r data-summary}
tse
```
