---
title: "Alpha Diversity"
author:
  - name: Muluh
params: 
  index: default
---

```{r variables}
# Alpha diversity
index <- params$index
variable      <- "diet" 
adjust.method <- "FDR"
test <- "Wilcoxon"
```

```{r data}
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

## Group-wise comparisons

-   **Index:** `r params$index`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

```{r run_diversity_tests}
# Calculate p-values
run_diversity_tests <- function(tse, comparisons, 
                                variable, index, 
                                adjust.method = "fdr", paired = FALSE) {
  # Get all values and groups
  values <- colData(tse)[[index]]
  groups <- colData(tse)[[variable]]
  
  # Create an empty results dataframe
  results <- data.frame(
    group1 = character(),
    group2 = character(),
    mean1 = numeric(),
    mean2 = numeric(),
    logFC = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through requested comparisons
  for (comp in comparisons) {
    group1 <- comp[1]
    group2 <- comp[2]
    
    # Subset data for just these two groups
    subset_indices <- which(groups %in% c(group1, group2))
    subset_values <- values[subset_indices]
    subset_groups <- groups[subset_indices]
    
    # Ensure paired data is properly aligned (if paired is TRUE)
    if (paired) {
      # Assuming 'subject' column exists for pairing
      subset_subjects <- tse$id[subset_indices]
      paired_data <- data.frame(
        subject = subset_subjects,
        group = subset_groups,
        value = subset_values
      ) %>%
        tidyr::pivot_wider(names_from = group, values_from = value) %>%
        drop_na()  # Drop pairs with missing data
      
      # Extract paired values
      paired_values <- paired_data %>% select(all_of(c(group1, group2)))
    }
    
    # Perform Wilcoxon test
    if (paired) {
      test_result <- wilcox.test(
        paired_values[[1]], paired_values[[2]],
        paired = TRUE, exact = FALSE
      )
    } else {
      test_result <- wilcox.test(
        subset_values ~ subset_groups,
        exact = FALSE
      )
    }
    
    # Calculate means
    mean1 <- mean(values[groups == group1])
    mean2 <- mean(values[groups == group2])

    if (paired) {
       # Note: when the values are paired, we can calculate FC separately
       # for every individual and summarize afterwards. 
       FC <- mean(values[groups == group2] / values[groups == group1])
    } else {
       FC <- mean2/mean1
    }

    # Add results
    results <- rbind(results, data.frame(
      group1 = group1,
      group2 = group2,
       mean1 = mean1,
       mean2 = mean2,
       logFC = log2(FC),
      p.value = test_result$p.value,
      stringsAsFactors = FALSE
    ))

  }
  
  # Add adjusted p-values using specified method
  # results$p.adj <- p.adjust(results$p.value, method = adjust.method)
  
  return(results)
}
```

```{r pvalues-before}
tse_before <- tse[, tse$visit == 1]
# Calculate statistical results
results_before <- run_diversity_tests(tse_before, comparisons_before, variable, index, adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse_before)[[index]][colData(tse_before)[[variable]] == group1[i]]
    vals2 <- colData(tse_before)[[index]][colData(tse_before)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
tse_after <- tse[, tse$visit == 2]
# Calculate statistical results
results_after <- run_diversity_tests(tse_after, comparisons_after, variable, index, adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse_after)[[index]][colData(tse_after)[[variable]] == group1[i]]
    vals2 <- colData(tse_after)[[index]][colData(tse_after)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r pvalues-paired}
diet_groups <- unique(tse$diet)
# Calculate statistical results
results_paired <- lapply(diet_groups, function(group) {
  # Subset tse for the current meal group
  tse_subset <- tse[, tse$diet == group]
  
  # Run the diversity tests
  res <- run_diversity_tests(
    tse = tse_subset, 
    comparisons = comparisons_paired, 
    variable = "timepoint", 
    index, 
    adjust.method = adjust.method, 
    paired = TRUE
  )
  
  # Add the meal group as a new column
  res$diet <- group
  
  return(res)
}) %>% 
  bind_rows()

results_paired <- results_paired[, c("diet", setdiff(names(results_paired), "diet"))]

# results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 1),
    mean2 = round(mean2, 1),
    logFC = round(logFC, 2),
    p.value = round(p.value, 3)
  ) %>%
  select(diet, group1, group2, mean1, mean2, logFC, p.value) %>% 
  arrange(diet)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[index]][colData(tse)[["timepoint"]] == group1[i]]
    vals2 <- colData(tse)[[index]][colData(tse)[["timepoint"]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```

```{r boxplot-before}
# Create a boxplot
p1 <- plotColData(
    tse_before, x = variable, y = index, colour_by = "timepoint",
    show_boxplot = TRUE, other_fields = c("timepoint"),
    show_violin = FALSE) +
    scale_color_manual(
        values = c("before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # Remove legend
        legend.position = "none"
    ) + coord_fixed(ratio = 0.5) +
    ggpubr::stat_pvalue_manual(
      data = final_results_before,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "Before treatment",
      x = "",
      y = paste("Alpha Diversity -", index)
     ) #+ scale_y_log10()
```

```{r boxplot-after}
# Create a boxplot
p2 <- plotColData(
    tse_after, x = variable, y = index, colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint")) + 
    scale_color_manual(
        values = c("after" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) + coord_fixed(ratio = 0.5) +
    ggpubr::stat_pvalue_manual(
      data = final_results_after,
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      label = "p.adj"
    )  +
    labs(
      title = "After treatment",
      x = "",
      y = paste("Alpha Diversity -", index)
    ) #+ scale_y_log10()
```

```{r all-groups-1v2}
p3 <- plotColData(
    tse, x = "timepoint", y = index, colour_by = "timepoint",
    show_boxplot = TRUE,
    show_violin = FALSE, other_fields = c("timepoint", "diet")) +
    scale_color_manual(
        values = c("after" = "black", "before" = "black"),
        name = "Timepoint"
    ) +
    theme(
        text = element_text(size = 10),
        # remove legend
        legend.position = "none"
    ) +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 3)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )  +
    labs(
      title = "Within group",
      x = "",
      y = paste("Alpha Diversity -", index)
    ) +
    facet_wrap(~diet, ncol = 4, scales = "free")
```

```{r changes1}
tse_new <- tse[,tse$paired == "yes"]
colData(tse_new) <- DataFrame(
    colData(tse_new) %>%
        as_tibble() %>%
        tidyr::drop_na({{ variable }}) %>%
        filter(paired == "yes") %>%
        group_by(id) %>%
        mutate(
            change = if_else(time == 2, .data[[index]] - .data[[index]][time == 1], 0),
            change = if_else(time == 1, change[time == 2], change)
        ) %>%
        ungroup() %>%
        as.data.frame(check.names = FALSE)
)

div_change <- plotColData(
    tse_new, x = "timepoint", y = index,
    show_boxplot = TRUE,
    show_violin = FALSE,
    other_fields = c("group", "id", "diet", "change")) + 
    facet_wrap(~diet, ncol = 4, scales="fixed", axis.labels = "margins") +
    theme(
        text = element_text(size = 10),
    ) +
    labs(
        title = "Paired alpha diversity",
        x = "",
        y = paste("Alpha Diversity -", index)
    ) + 
    geom_line(
        aes(group = id, color = change)
    ) +
    scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab") +
    #scale_y_log10() +
    geom_text(
        data = final_results_paired,
        aes(
            x = 1.5,                 
            y = y.position,          
            label = paste0("p = ", signif(p.value, 3)),
            group = diet        
        ),
        inherit.aes = FALSE,     
        size = 3,                 
        color = "black"            
    )

```

```{r visualize-plot}
#| fig-width: 5
#| fig-height: 4
#| fig.align: "center"
y_min <- min(c(tse_before[[index]], tse_after[[index]]), na.rm = TRUE)
y_max <- max(c(tse_before[[index]], tse_after[[index]]), na.rm = TRUE)

# Get the maximum y-position from your annotations
max_annotation_y <- max(c(
    final_results_before$y.position, 
    final_results_after$y.position
), na.rm = TRUE)

# Use the larger of data max and annotation position for the upper limit
y_max <- max(y_max, max_annotation_y)

y_range <- y_max - y_min
y_max <- y_max + (y_range * 0.2)  

# Apply to both plots
p1 <- p1 + coord_cartesian(ylim = c(y_min, y_max))
p2 <- p2 + coord_cartesian(ylim = c(y_min, y_max))
wrap_plots(p1, p2, axes = "collect")
```

```{r Explanations}
before_exp_sh <- "The alpha diversity between the 'oat' and 'rice' groups appears to be very similar, with minimal difference in the mean values (both 4.00). 

The log2FC indicates almost no change between the two groups in terms of diversity `(0.02)`, further supporting the notion of no substantial difference.

The p.adj of '0.373' shows that the difference between the groups is not statistically significant. Thus, we would conclude that the treatment ('oat' and 'rice') does not have a significant impact on the alpha diversity of the groups at the baseline."

after_exp_sh <- "After treatment, the mean alpha diversity for 'oat' is `3.90`, and for 'rice' is `4.10`, showing a slight difference between the two groups, with 'rice' having slightly higher diversity.\n

The log2FC value of `0.06` indicates a very small positive difference, suggesting that the 'rice' group experienced a slight increase in diversity compared to 'oat'.

The p.adj of `0.0480` indicates that the difference between the groups is statistically significant (since it's less than `0.05`), meaning that the observed difference in diversity between the 'oat' and 'rice' groups after treatment is likely not random."

paired_exp_sh <- "For the oat diet, there was a slight decrease in alpha diversity from Timepoint 1 `(mean = 4.00) `to Timepoint 2 `(mean = 3.90)`, but the change was not statistically significant `(p-value = 0.151)`.

For the rice diet, there was a slight increase in alpha diversity from Timepoint 1 `(mean = 4.00)` to Timepoint 2 `(mean = 4.10)`, but again, the change was not statistically significant `(p-value = 0.0780)`."

lmm_sh <- "
(Intercept):

Estimate: `4.04`
This represents the baseline value of the alpha diversity when both the diet and timepoint are at their reference categories (oat for diet and before for timepoint). `Pr(>|t|): 4.19e-84`, which is extremely small and indicates highly significant evidence that the intercept term is different from zero.

dietrice:

Estimate: `-0.08`
This is the difference in alpha diversity between rice and the reference diet (oat), after adjusting for the effect of time. `Pr(>|t|): 0.545` suggests that the diet (rice vs oat) effect is not statistically significant. The difference in alpha diversity between rice and oat diets is not substantial enough to conclude that it is different.

time:

Estimate: `-0.06`
This is the difference in alpha diversity between timepoints ( after vs before), after adjusting for the effect of diet. `Pr(>|t|): 0.148` indicates that the effect of time (before vs after) on alpha diversity is not statistically significant. This suggests that alpha diversity does not change significantly between timepoints in the study.

dietrice:time (Interaction ):

Estimate:` 0.13`
This is the interaction between diet (rice vs oat) and timepoint (before vs after). The positive estimate suggests that the effect of time on alpha diversity differs between the rice and oat diets. `Pr(>|t|): 0.0378` indicates that the interaction between diet and timepoint is statistically significant at the usual significance level `(0.05)`. This suggests that the relationship between alpha diversity and timepoint differs for the rice and oat diets.
"
before_exp_obs <- "Before treatment, the rice group had a slightly higher observed alpha diversity `(231.30)` compared to the oat group `(210.90)`. The fold change (log2FC) of `0.13` suggests a small increase in richness for the rice group. However, the corrected p-value `(p.adj = 0.133)` indicates that this difference is not statistically significant, meaning it could be due to random variation rather than a true effect of diet."

after_exp_obs <- "

After treatment, the rice group `(233.00)` continued to show a higher observed alpha diversity compared to the oat group `(206.90)`, with a fold change (log2FC) of `0.17`. The adjusted p-value `(p.adj = 0.0540)` is approaching statistical significance, suggesting a potential effect of diet on diversity, though it remains marginal."

paired_exp_obs <- "

Oat group: Observed alpha diversity slightly decreased from `210.90` to `206.90` (log2FC = `0.04`, p = `0.120`), indicating a small but non-significant reduction over time.

Rice group: Diversity increased modestly from `231.30` to `233.00` (log2FC = `0.13`, p = `0.363`), but this change is also not statistically significant.

Overall, there were no significant within-group changes in observed alpha diversity over time, suggesting that dietary intervention had a minimal effect on diversity shifts within each group."

lmm_obs <- "

Diet effect (rice vs. oat): No significant difference between rice and oat groups `(Estimate = 0.12, p = 0.333`), suggesting baseline diversity was similar.

Time effect: A slight decrease in diversity over time `(Estimate = -0.03, p = 0.177)`, but this was not statistically significant.

Diet × Time interaction: The interaction term `(Estimate = 0.04, p = 0.201)` suggests that the change in diversity over time did not significantly differ between the oat and rice groups."

if (index == "observed") {
  before_exp <- before_exp_obs
  after_exp <- after_exp_obs
  paired_exp <- paired_exp_obs
  lmm_exp <- lmm_obs
} else {
  before_exp <- before_exp_sh
  after_exp <- after_exp_sh
  paired_exp <- paired_exp_sh
  lmm_exp <- lmm_sh
}
```


### Table of the top findings

#### Across-group

##### Before treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-before}
#| label: table-of-findings-before
if (knitr::is_html_output()) {
  datatable(
      final_results_before %>% select(-y.position), 
      options = list(
        pageLength = 6,
        dom = 'Bfrtip'
      ),
      caption = "Table of alpha diversity comparisons before Treatment",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_before %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity comparisons before Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

##### After treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**


```{r table-of-findings-after}
#| label: table-of-findings-after
if (knitr::is_html_output()) {
  datatable(
    final_results_after %>% select(-y.position), 
    options = list(
      pageLength = 6,
      dom = 'Bfrtip'
    ),
    caption = "Table of alpha diversity comparisons after Treatment",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_after %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity comparisons after Treatment",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

```{r viz-div_change}
#| fig-width: 5
#| fig-height: 3
#| fig.align: "center"
div_change
```

#### Within-group

##### Paired treatment

-   Multiple testing correction: **paired** -none
-   Statistical test: **`r test`**

```{r table-of-findings-paired}
#| label: table-of-findings-paired

if (knitr::is_html_output()) {
  datatable(
    final_results_paired %>% select(-y.position), 
    options = list(
      pageLength = 4,
      dom = 'Bfrtip'
    ),
    caption = "Table of alpha diversity paired between timepoint(1 and 2)",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.value", digits = 3)
} else {
  library(kableExtra)
  final_results_paired %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of alpha diversity paired between timepoint(1 and 2)",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

### LMM

```{r lmer-stuffm}
adjust.method <- "FDR"
test <- "paired Wilcoxon"
```

```{r lmer-stuff}
formula_used <- "~ diet * intervention + (1 | id)"

if (index == "observed") {
  tse$observed_log2 <- log2(tse$observed)
  res <- run_lmer(tse, "observed_log2")
} else {
  res <- run_lmer(tse, index)
}
```

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**
-   Formula used: **`r formula_used`**

```{r table-fixed-fx}
sjPlot::plot_model(res)
sjPlot::tab_model(res)
```

## Alternate analyses

```{r stratified-timepoint}
# First analysis - stratified by time point
tbl_a_s <- colData(tse) %>% as_tibble() %>%
    select(c("timepoint", index, "meal_group", "age")) %>%
    tbl_strata(strata = timepoint, .combine_with = "tbl_merge",
               .tbl_fun = ~.x %>% tidyr::drop_na() %>% 
                   lm(formula = as.formula(
                       paste0(index, " ~ meal_group + age")),
                      data = .) %>% 
                   tbl_regression(
                       intercept = TRUE, 
                       include = everything()) %>%
                   add_global_p(quite=TRUE, keep=TRUE))
```

```{r paired}
# Paired analysis
tbl_a_s_diff <- colData(tse) %>% as_tibble() %>%
    select(c("timepoint", index, variable, "meal_group", "paired", "age")) %>% 
    tidyr::drop_na(variable) %>% filter(paired=="yes") %>% 
    lmerTest::lmer(
        formula = as.formula(
              paste0(index, " ~ meal_group + age + (1|", variable, ")")
        ),
        data = .) %>% 
    tbl_regression(
        intercept = TRUE, 
        include = everything()) %>%
    add_global_p(quite=TRUE, keep=TRUE) %>% 
    modify_header(label = "**Paired test between timepoints (1 and 2)**")
```

-   **Beta (Estimate)**: Shows the effect size and direction of a predictor on the outcome. Positive means an increase, negative a decrease.

-   **95% CI**: Range where the true beta is likely to fall, reflecting precision.

-   **p-Value**: Tests if beta differs from zero; p\<0.05p \< 0.05p\<0.05 suggests significance.

```{r table-res1}
knitr::knit_print(tbl_a_s)
```

```{r table-res2}
knitr::knit_print(tbl_a_s_diff)
```

```{r Plotting-code}
# Plotting code
# Set up reference levels
ref <- levels(factor(colData(tse)[["meal_group"]]))[1]
lvs <- levels(factor(colData(tse)[["meal_group"]]))[-1]
comps <- lapply(lvs, function(lv) c(ref,lv))
```

```{r first-plot}
# First plot
p1 <- colData(tse) %>% as_tibble() %>%  tidyr::drop_na(variable) %>%
    as.data.frame(check.names=FALSE) %>%
    ggplot(aes_string(x="meal_group", y = index))+
    geom_boxplot(outlier.shape = NA)+
    geom_jitter(width=0.1, size=0.5)+
    facet_wrap(~timepoint, ncol = 2, scales="free")+
    geom_signif(comparisons = comps, map_signif_level = FALSE,
                step_increase=0.2, textsize = 2.5)+
    labs(x="", y = paste(index, "index"))

# Process p-values for plot
p_tmp <- ggplot_build(p1)
df_1 <- p_tmp$data[[3]] %>% mutate(label=gsub(paste0(ref,"|\\-|[0-9]"),"",
                                              .data[["group"]]))
df_2 <- tbl_a_s$table_body %>% filter(label %in% lvs) %>%
    select(c(label, contains("p.value"))) %>% 
    tidyr::pivot_longer(cols = contains("p.value"),
                        names_to="PANEL", names_prefix="p.value_") %>% 
    mutate(PANEL=factor(PANEL, levels = levels(df_1$PANEL)))
p_tmp$data[[3]] <- merge(df_1,df_2, by = c("PANEL","label")) %>%
    mutate(annotation=round(value,3)) %>% select(colnames(p_tmp$data[[3]]))
p1 <- p_tmp %>% ggplot_gtable() %>% ggplotify::as.ggplot()


```

```{r second-plot}
# Second plot
df_pval <- tbl_a_s_diff$table_body %>% filter(label %in% lvs) %>%
    select(c(label, contains("p.value")))
pos_j <- position_jitter(width = 0.1, height = 0)
p2 <- colData(tse) %>% as_tibble() %>%  tidyr::drop_na({{ variable }}) %>%
    filter(paired=="yes") %>% group_by(.data[[variable]]) %>% arrange(id, timepoint) %>%
    mutate(
        change = .data[[index]] - dplyr::first(.data[[index]])
    ) %>% tidyr::fill(change, .direction = "downup") %>%  
    as.data.frame(check.names=FALSE) %>%
    ggplot(aes(x = timepoint, y = .data[[index]]))+
    geom_boxplot(outlier.shape = NA)+
    geom_point(position = pos_j, size = 0.5) + 
    geom_line(aes(group = id, color = change), position = pos_j) +
    scale_color_gradient2(midpoint=0, low="blue", mid="white", high="red", space ="Lab")+
    facet_wrap(as.formula(paste0("~", "meal_group")), nrow = 1,
               labeller = function(df) {
                   df[,1] <- sapply(df[,1], function(lb){
                       if (lb!=ref) {
                           lb <- paste0(lb, " (p-val = ",
                                        df_pval %>% filter(label==lb) %>%
                                            select(contains("p.value")) %>%
                                            .[[1]] %>% round(3)
                                        ,")")
                           lb
                       } else {
                           as.character(lb)
                       }
                   }) %>% unlist()
                   df
               })+
    labs(x="", y = paste(index, "index"))


```

## Data Summary

```{r data-summary}
tse
```
