---
title: "Correlation analysis"
author: ""
params: 
  tax.level: default
---

```{r setup}
# load packages
library(mia)
library(dplyr)
library(DT)
library(shadowtext)
library(ComplexHeatmap)
# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse_delta.Rds")

# Define variables and formulas
prevalence.th <- 10/100
detection.th <- 0.1/100
# Other taxonomic levels can be checked with altExpNames(tse)
```

# Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%
-   Number of taxonomic groups: `r nrow(tse)`
-   Transformation: _clr_

```{r}
# Get the variables to be correlated
# These are computed differences of the vvariables. 
scfa_vars <- c("acetic_delta", "propionic_delta", "butyric_delta", 
                 "isobutyr_delta", "succinic_delta",
                 "valeric_delta", "isovaler_delta", "lactic_delta", 
                 "indolelactic_delta", "indolebutyric_delta",
                 "indolepropionic_delta", "heptanoic_delta")
  
  inflamm_markers <- c("IL1B_delta", "MMP12_delta", 
                       "TNFSF12_delta", "EGF_delta")
  
  diet_vars <- c("fat_delta", "carb_delta", "prot_delta", "fiber_delta", 
                 "SFA_delta", "MUFA_delta", "PUFA_delta")
  diet_vars <- c(
  # Macronutrients (absolute)
  "fat_delta", "carb_delta", "prot_delta", "fiber_delta",
  # Fatty acids (absolute)
  "SFA_delta", "MUFA_delta", "PUFA_delta",
  # Macronutrient percentages
  "fat_p_delta", "carb_p_delta", "prot_p_delta",
  # Fatty acid percentages
  "SFA_p_delta", "MUFA_p_delta", "PUFA_p_delta",
  # Fiber subtype
  "fiber_e_delta",
  # Micronutrients
  "Mg_delta", "Fe_delta", "Zn_delta"
)
  
  bio_vars <- c("waist_delta", "Kol_delta", "LDLkol_delta", 
                "HDLkol_delta", "Trigly_delta")
```
## Calculate correlations
```{r}
all_vars <- list(scfa_vars = c(scfa_vars), 
                 inflamm_markers = c(inflamm_markers), 
                 diet_vars = c(diet_vars), 
                 bio_vars = c(bio_vars))
cross_assoc_results <- list()

# Run getCrossAssociation for each variable and store the full result
results_list <- lapply(all_vars, function(some_var) {
getCrossAssociation(
    tse, 
    assay.type1 = "clr_delta", 
    mode = "matrix",
    col.var2 = some_var,  
    method = "kendall", 
    p_adj_method = "fdr",
    test.signif = TRUE, 
    show.warnings = FALSE, 
    verbose = FALSE
  )
})
```
## Plot heatmaps
```{r}
## healper to create heatmaps
create_heatmap <- function(res) {
      add_signif <- function(j, i, x, y, width, height, fill) {
        # If the p-value is under threshold
        if( !is.na(res$p_adj[i, j]) & res$p_adj[i, j] < 0.05 ){
            # Print "X"
            grid.shadowtext(
                sprintf("%s", "X"), x, y, gp = gpar(fontsize = 8, col = "#f5f5f5"))
        }
    }
  
  Heatmap(
    res$cor,
    name = "correlation",
    cell_fun = add_signif,
    heatmap_legend_param = list(title = "Correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
  )
}
```

## microbiota vs. Short-chain fatty acids (12 SCFAs in total)
```{r}
heatmap_scfa <- create_heatmap(results_list$scfa_vars)
heatmap_scfa
```

## microbiota vs. significant inflammation markers (IL-1B, MMP12, TNSF-12, EGF)
```{r}
heatmap_inflamm_markers <- create_heatmap(results_list$inflamm_markers)
heatmap_inflamm_markers
```

## ## microbiota vs. dietary intake (total fat, carbohydrates, protein, fiber, SFA, MUFA, PUFA)
```{r}
heatmap_diet_vars <- create_heatmap(results_list$diet_vars)
heatmap_diet_vars
```

## microbiota vs. biochem/antrop. (waist circumference, total cholesterol, LDL, HDL, Triglycerides)
```{r}
heatmap_bio_vars <- create_heatmap(results_list$bio_vars)
heatmap_bio_vars
```


