---
title: "Correlation analysis"
author: ""
params: 
  tax.level: default
---

```{r setup}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse_delta.Rds")

# Define variables and formulas
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- params$tax.level

if (tax.level == "phylum_prevalent") {
    tse <- tse
} else {
    tse <- altExp(tse, tax.level)
}
```


# Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%
-   Number of taxonomic groups: `r nrow(tse)`
-   Transformation: _clr_

```{r}
scfa_vars <- c("acetic_delta", "propionic_delta", "butyric_delta", 
                 "isobutyr_delta", "succinic_delta",
                 "valeric_delta", "isovaler_delta", "lactic_delta", 
                 "indolelactic_delta", "indolebutyric_delta",
                 "indolepropionic_delta", "heptanoic_delta")
  
  inflamm_markers <- c("IL1B_delta", "MMP12_delta", 
                       "TNFSF12_delta", "EGF_delta")
  
  diet_vars <- c("fat_delta", "carb_delta", "prot_delta", "fiber_delta", 
                 "SFA_delta", "MUFA_delta", "PUFA_delta")
  
  bio_vars <- c("waist_delta", "Kol_delta", "LDLkol_delta", 
                "HDLkol_delta", "Trigly_delta")
```


## microbiota vs. Short-chain fatty acids (12 SCFAs in total)
### Table

```{r scfa_sum}
#| label: tbl-table-of-scfa_sum
result1 <- getCrossAssociation(
    tse, assay.type1 = "clr_delta",
    col.var2 = scfa_vars,
    test.signif = TRUE, show.warnings = FALSE, verbose = FALSE) 

result <- result1 %>%
    select(-pval) %>%
    dplyr::rename(SCFAs = "Var2", microbes = "Var1")

datatable(
  result,
  options = list(pageLength = 10, dom = 'Bfrtip'),
  caption = "Correlation table of Short-chain fatty acids",
  rownames = FALSE
)  %>%
      formatSignif(columns = c("cor", "p_adj"), digits = 3)
```

### Heatmap

```{r}
#| label: fig-figure-heat-scfa
#| fig-width: 8
#| fig-height: 8
result <- result %>%
  mutate(
    signif_label = case_when(
      p_adj < 0.05 ~ "*",
      p_adj < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(cor), paste0(round(cor, 4), signif_label), "")
  )

result <- result[result$p_adj < 0.804, ]

# Plot
ggplot(result, aes(x = SCFAs, y = microbes, fill = cor)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Correlation"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

### Scatterplot

```{r}

top_bottom_corr <- result1 %>%
  arrange(desc(cor)) %>%
  slice_head(n = 5) %>%
  bind_rows(
    result1 %>%
      arrange(cor) %>%
      slice_head(n = 5)
  )

ggplot(top_bottom_corr, aes(x = cor, y = reorder(paste(Var1, Var2, sep = " vs "), cor))) +
  geom_point(aes(color = cor > 0), size = 3) +
  labs(
    title = "Top 5 and Bottom 5 Correlations",
    x = "Correlation",
    y = "Microbe vs Variable"
  ) +
  scale_color_manual(values = c("red", "blue"), labels = c("Positive", "Negative")) +
  theme_minimal()
```


## microbiota vs. significant inflammation markers (IL-1B, MMP12, TNSF-12, EGF)
### Table

```{r inflamm_sum}
#| label: tbl-table-of-inflamm_sum
result2 <- getCrossAssociation(
    tse, assay.type1 = "clr_delta",
    col.var2 = inflamm_markers,
    test.signif = TRUE, show.warnings = FALSE, verbose = FALSE)
result <- result2 %>%
    select(-pval) %>%
    dplyr::rename(`Inflammation markers` = "Var2", microbes = "Var1")
datatable(
  result,
  options = list(pageLength = 10, dom = 'Bfrtip'),
  caption = "Correlation table of significant inflammation markers",
  rownames = FALSE
)  %>%
      formatSignif(columns = c("cor", "p_adj"), digits = 3)
```

### Heatmap

```{r}
#| label: fig-figure-heat-inf
#| fig-width: 8
#| fig-height: 8
result <- result %>%
  mutate(
    signif_label = case_when(
      p_adj < 0.05 ~ "*",
      p_adj < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(cor), paste0(round(cor, 4), signif_label), "")
  )

result <- result[result$p_adj < 0.7, ]

# Plot
ggplot(result, aes(x = `Inflammation markers`, y = microbes, fill = cor)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Correlation"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

### Scatterplot

```{r}
top_bottom_corr <- result2 %>%
  arrange(desc(cor)) %>%
  slice_head(n = 5) %>%
  bind_rows(
    result2 %>%
      arrange(cor) %>%
      slice_head(n = 5)
  )

ggplot(top_bottom_corr, aes(x = cor, y = reorder(paste(Var1, Var2, sep = " vs "), cor))) +
  geom_point(aes(color = cor > 0), size = 3) +
  labs(
    title = "Top 5 and Bottom 5 Correlations",
    x = "Correlation",
    y = "Microbe vs Variable"
  ) +
  scale_color_manual(values = c("red", "blue"), labels = c("Positive", "Negative")) +
  theme_minimal()
```

## microbiota vs. dietary intake (total fat, carbohydrates, protein, fiber, SFA, MUFA, PUFA)
### Table

```{r diet_sum}
#| label: tbl-table-of-diet_sum
result3 <- getCrossAssociation(
    tse, assay.type1 = "clr_delta",
    col.var2 = diet_vars,
    test.signif = TRUE, show.warnings = FALSE, verbose = FALSE)
result <- result3 %>%
    select(-pval) %>%
    dplyr::rename(`Dietary fats` = "Var2", microbes = "Var1")
datatable(
  result,
  options = list(pageLength = 10, dom = 'Bfrtip'),
  caption = "Correlation table of dietary intake",
  rownames = FALSE
)  %>%
      formatSignif(columns = c("cor", "p_adj"), digits = 3)
```

### Heatmap

```{r}
#| label: fig-figure-heat-diet
#| fig-width: 8
#| fig-height: 8
result <- result %>%
  mutate(
    signif_label = case_when(
      p_adj < 0.05 ~ "*",
      p_adj < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(cor), paste0(round(cor, 4), signif_label), "")
  )

result <- result[result$p_adj < 0.7, ]

# Plot
ggplot(result, aes(x = `Dietary fats`, y = microbes, fill = cor)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Correlation"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

### Scatterplot

```{r}
top_bottom_corr <- result3 %>%
  arrange(desc(cor)) %>%
  slice_head(n = 5) %>%
  bind_rows(
    result3 %>%
      arrange(cor) %>%
      slice_head(n = 5)
  )

ggplot(top_bottom_corr, aes(x = cor, y = reorder(paste(Var1, Var2, sep = " vs "), cor))) +
  geom_point(aes(color = cor > 0), size = 3) +
  labs(
    title = "Top 5 and Bottom 5 Correlations",
    x = "Correlation",
    y = "Microbe vs Variable"
  ) +
  scale_color_manual(values = c("red", "blue"), labels = c("Positive", "Negative")) +
  theme_minimal()
```

## microbiota vs. biochem/antrop. (waist circumference, total cholesterol, LDL, HDL, Triglycerides)
### Table

```{r bio_sum}
#| label: tbl-table-of-bio_sum
result4 <- getCrossAssociation(
    tse, assay.type1 = "clr_delta",
    col.var2 = bio_vars,
    test.signif = TRUE, show.warnings = FALSE, verbose = FALSE)
result <- result4 %>%
    select(-pval) %>%
    dplyr::rename(`biochem/antrop` = "Var2", microbes = "Var1")
datatable(
  result,
  options = list(pageLength = 10, dom = 'Bfrtip'),
  caption = "Correlation table of biochem/antrop",
  rownames = FALSE
)  %>%
      formatSignif(columns = c("cor", "p_adj"), digits = 3)
```

### Heatmap

```{r}
#| label: fig-figure-heat-biochem
#| fig-width: 8
#| fig-height: 8
result <- result %>%
  mutate(
    signif_label = case_when(
      p_adj < 0.05 ~ "*",
      p_adj < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(cor), paste0(round(cor, 4), signif_label), "")
  )

result <- result[result$p_adj < 0.6, ]

# Plot
ggplot(result, aes(x = `biochem/antrop`, y = microbes, fill = cor)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Correlation"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

### Scatterplot

```{r}

top_bottom_corr <- result4 %>%
  arrange(desc(cor)) %>%
  slice_head(n = 5) %>%
  bind_rows(
    result4 %>%
      arrange(cor) %>%
      slice_head(n = 5)
  )

ggplot(top_bottom_corr, aes(x = cor, y = reorder(paste(Var1, Var2, sep = " vs "), cor))) +
  geom_point(aes(color = cor > 0), size = 3) +
  labs(
    title = "Top 5 and Bottom 5 Correlations",
    x = "Correlation",
    y = "Microbe vs Variable"
  ) +
  scale_color_manual(values = c("red", "blue"), labels = c("Positive", "Negative")) +
  theme_minimal()
```